---
title: "430_project"
author: "Marina Tornorsam"
output: 
  html_document:
    toc: true
  github_document: 
    toc: true
---

## FRE 490 Spatial Data Analysis and Remote Sensing

##### Instructor: Jonathan Proctor

##### Teaching Assistant: Bhavita Ragoo

### Final Project workflow

-   This project entails: 7-minute presentation,
-   written analysis documenting the project and the results (5-pages double spaced text plus figures, brief intro/research question/methods section/results

### Objectives:

In this project, I will design a research question and then conduct the analysis and write up the results. Initial questions (based on proposal):

1)  How does air quality vary across different land use types in the lower Mekong region?

2)  Does air quality disproportionately impact vulnerable/ high population density areas?

#### *Refined question:*

1.  *To what extent do seasonal fire hotspots (VIIRS) correlate with specific land use types in Cambodia?*
    1.  *What is the spatial and temporal distribution of fire hotspots across land use types in Cambodia (2019–2025)?*
    2.  *Has the timing and intensity of the fire season shifted over the period, and is this shift associated with climate variability (ENSO/precipitation anomalies)?*
    3.  *Are fire hotspots disproportionately concentrated in edge/transitional zones between forest and agricultural land use types?*
2.  *To what extent is there influence from human interaction on fire incidence and occurrence?*
    1.  *Are provinces and communes with higher human footprint index associated with greater fire frequency and intensity?*
    2.  *Does fire incidence differ significantly inside versus outside protected areas, and does this vary by land use type?*

side quest: What are the implications of air quality and downwind potentially vulnerable communities? Which populated areas are at an elevated exposure risk from fire smoke based on prevailing wind direction and proximity to high-FRP fire clusters (based on 2025?)

### Hypothesis

Fire hotspot density and intensity in Cambodia are disproportionately concentrated in agricultural and forest-agricultural transition land use types (like savanna and shrubland) with the fire season showing a detectable temporal shift for start date and end date. Fire desnity and intensity are positively associated with human footprint index at the province level suggesting that anthropogenic land management is a primary driver of fire incidence and spatial distribution.

Null hypothesis: Fire hotspot density and intensity shows no significant association with land use type, human footprint index, and protected areas indicating that the fire incidence is driven by natural climate variability.

Areas with higher fire density occurs in agricultural land use types and in transitional areas (proximate to cropland and forested areas). Increased fire activity occurs in areas with high human footprint index.

### Datasets

1)  Active fire points: NASA VIIRS (375m resolution): 2012-present; thermal anomalies [link](https://firms.modaps.eosdis.nasa.gov/map/)

    1)  FIRMS -\> GEE: <https://code.earthengine.google.com/?scriptPath=Examples%3ADatasets%2FFIRMS%2FFIRMS>

    2)  Considering using Burned area with MODIS -\> GEE: <https://code.earthengine.google.com/?scriptPath=Examples%3ADatasets%2FMODIS%2FMODIS_061_MCD64A1>

2)  Land use: Regional land cover monitoring system (RLCMS) by SERVIR-SEA (30m resolution) (2012 to 2023) [link](https://landcovermapping.org/en/about/) (Saah et al 2020)

    1)  GEE: <https://code.earthengine.google.com/ddc999619aa0c29514cccfc673a4f7e1>

3)  Cambodian spatial distribution of population and age and gender structures from WorldPop (100m) (2015-2020) [link](https://data.humdata.org/organization/worldpop?groups=khm&q=&ext_page_size=25)

    1)  Estimated population GEE: <https://code.earthengine.google.com/25af8a8783b214b89b8f5bef55e1be80>

4)  Cambodia administrative boundaries: Latest data received 2024 [link](https://data.humdata.org/dataset/cod-ab-khm)

    1)  Natural protected areas 1993-2023: [link](https://data.opendevelopmentmekong.net/en/dataset/protectedareas) national park, wildlife sanctuary, protected landscape, multiple use, natural heritage site, Ramsar site, biosphere reserve, marine national park

5)  Human footprint index: relative human influence in each terrestrial biome (%); updated to 2025; 1 km resolution created from human population density, human land use and infrastructure (built up areas, nighttime lights, land use/land cover), human access; least influence - 0%, most influenced - 100%; WCS/Columbia University [link](https://hub.arcgis.com/maps/65518e782be04e7db31de65d53d591a9/about)

    1)  GEE: (Kennedy et al., 2019) <https://code.earthengine.google.com/80e4e1006b6ec26ece0975766adc57b1>

Optional/ potential datasets:

1)  OpenAQ: ground validation for air pollution detected from local devices [link](https://explore.openaq.org/#3.63/12.44/99.08)

2)  Poverty CHand vulnerability map for Cambodia: AI-calculated using dimensions that include education, health, living standard, employment, monetary; launched in 2024 [link](https://cambodiapovertymapping.sig-gis.com/en/home/)

3)  CHIRPS Precipitation daily [link](https://code.earthengine.google.com/?scriptPath=Examples%3ADatasets%2FUCSB-CHG%2FUCSB-CHG_CHIRPS_DAILY)

    Climate hazards center infrared precipitation with station data. 30+ years; 0.5 degree resolution satellite imagery; 1981-2026

4)  ERA5 daily aggregates - combines model data with observations across the world; Copernicus Climate Change Service; air temperature, dewpoint temperature, total precipitation, mean sea level pressure, surface pressure, u-component of wind and v-component of wind

### Analysis

Fire hotspot density per commune

land cover/ use % per communerasterize the fire hot spot data into one pixel –\> converts individual points into a grid –\> pixel to pixel comparison

calculate the maximum FRP by sum daily 1 km –\> proxy for fire intensity –\> compare to fire counts for burned area

Characterizing the temporal progression of agricultural fires (see Liu et al 2021)

correlation

raster::overlay with population density and human footprint index –\> regression analysis

1.  Cross sectional modeling
2.  cross validation regression
3.  moving window regression
4.  panel modeling

### Limitations

Resolutions of datasets

Land cover land use datasets are only up to date from SERVIR to 2023, thus, I applied 2023 LULC to fire data from 2024, 2025

### Future Work

This preliminary analysis aims to incorporate on-the-ground data validation and novel insights on land use, hidden social-ecological values, and human-interaction landscape dynamics with an emphasis on fire use and livelihoods. This data collection will use the [LiteFarm](https://r9-kdd3.math.ubc.ca/home/) application.

## Setup

Install and load necessary packages:

```{r}
rm(list = ls())
library(raster) # handle raster data
library(ggplot2) #data visulations
library(ggthemes) #enhance data vis
library(stargazer) # regression outputs
library(cowplot) # putting plots together
library(sf) # handling vector operations - spatial features
library(dplyr) # data wrangling
library(scales) #graphical scales for vis
library(tidyr) # data wrangling
library(lubridate) #parsing dates
library(terra) #handle raster data (faster)
library(nngeo) # k-nearest neighbor join for spatial data 
library(patchwork) # combine plots into one exportable image
library(stringr) # simple, consistent wrappers for common string operations
library(kableExtra) # pretty tables
library(broom) # convert statistical objects into tidy tables
library(GWmodel) # geographic weighted regression model
library(plm) # panel regression
library(tidyterra) # ggplot for rasters

setwd("C:/Users/marin/OneDrive - UBC/2. Coursework/2025 Term 2 - FRE 430/FRE 430/Final_project")
```

## Preliminary Analysis

### Fire activity

Standardize the data sets to the same coordinate system: WGS 84 / UTM zone 48N (EPSG:32648)

Temporal window: Fire season from January 1st to April 30th for years: 2019 to 2025 using VIIRS

```{r}
#focus on a specific folder with the dataset
file_list_fires <- list.files(path = "C:/Users/marin/OneDrive - UBC/2. Coursework/2025 Term 2 - FRE 430/FRE 430/Final_project/Cambodia_dl-03.01_viirs", 
                        pattern = "\\.shp$", 
                        full.names = TRUE)
# Read in all the files at one time using lapply
lapply(file_list_fires, function(x) names(st_read(x, quiet = TRUE)))
# read the files in as sf objects
sf_list <- lapply(file_list_fires, read_sf)
# combine all the years (2018 April - 2026 Jan) into one dataframe
all_fires <- bind_rows(sf_list)


# filter the dataframe for the fire season of Jan-April
all_fires <- all_fires %>%
  mutate(ACQ_DATE = as.Date(ACQ_DATE), # Convert character/factor to Date
         Month = month(ACQ_DATE),     # Extract month as a number (1-12)
         Year = year(ACQ_DATE))  # Extract year as a number) %>%

# create variable filtered for only fire season 
all_fires_jan_apr <- all_fires %>%
  filter(Year != 2018 & Year != 2026) %>%
  filter(Month >= 1 & Month <= 4) 
#calculate the count of fires for each year during the fire season
fire_season_counts <- all_fires_jan_apr %>%
  mutate(Year = year(ACQ_DATE)) %>%
  group_by(Year) %>%
  summarize(season_fires = n())

head(fire_season_counts)


```

```{r}
# #plot the fires for each year
# ggplot() +
# geom_sf(data = all_fires_jan_apr, aes(color = FRP), 
#         size = 0.5, alpha = 0.7) +
#   # grid by year
#   facet_wrap(~ Year, ncol = 4) +  
#   scale_color_viridis_c(option = "inferno", name = "FRP") +
#   theme_void() +
#   labs(title = "VIIRS Fire intensity in Cambodia (Jan-Apr)",
#        caption = "Source: FIRMS VIIRS 375m") +
#   coord_sf()
```

### Admin boundaries layer: provinces and communes

```{r}

# using the latest data for administrative boundaries - there are three levels - province, district, commune

file_list_admin <- list.files(path = "C:/Users/marin/OneDrive - UBC/2. Coursework/2025 Term 2 - FRE 430/FRE 430/Final_project/Cambodia_admin", 
                        pattern = "\\.shp$", 
                        full.names = TRUE)
# Read in all the files at one time using lapply
lapply(file_list_admin, function(x) names(st_read(x, quiet = TRUE)))
# read the files in as sf objects
sf_list_admin <- lapply(file_list_admin, read_sf)
# rename each file for the specific administrative boundary
country <- sf_list_admin[[1]]
provinces <- sf_list_admin[[2]]
districts <- sf_list_admin[[3]]
communes  <- sf_list_admin[[4]]
# simple plot for the provinces
ggplot()+
  geom_sf(data = provinces, fill = "gray95", color = "black") +
  theme_void() +
  labs(title = "Provinces of Cambodia")
# plot the communes
ggplot()+
  geom_sf(data = communes, fill = "gray95", color = "black") +
  theme_void() +
  labs(title = "Communes of Cambodia")
# count communes per province
commune_per_province <- communes %>%
  st_drop_geometry() %>%
  group_by(adm1_name) %>%
  summarise(num_communes = n()) %>%
  arrange(desc(num_communes))

print(commune_per_province)


```

### Provincial analysis: fire activity

-   Hotspot count: measuring frequency

-   mean FRP: identifying high intensity fires vs. smaller fires

-   Fire density: hotspots per km2: normalizes the data across land area

```{r}
# confirm that the datasets are using the same CRS
if (st_crs(all_fires_jan_apr) != st_crs(provinces)) {
  warning("CRS mismatch detected! Reprojecting fires to match provinces...")
  all_fires_jan_apr <- st_transform(all_fires_jan_apr, st_crs(provinces))
} else {
  message("CRS is in alignment!")
}
#repair administrative boundaries to ensure no overlapping boundaries
provinces <- st_make_valid(provinces)

# spatial join to ensure that each fire point is attributed to a province where it is located and filter out only the high confidence hotspots
fires_in_prov <- st_join(all_fires_jan_apr%>% filter(CONFIDENCE >= 80), provinces)

# aggregate by year and province 
# calculate total count, mean FRP, and fire density
prov_fire_stats <- fires_in_prov %>%
  st_drop_geometry() %>%    # Dropping geometry makes the calculation 100x faster
  group_by(Year, adm1_name, area_sqkm) %>% 
  summarize(
    total_hotspots = n(),
    avg_frp = mean(FRP, na.rm = TRUE),
    max_frp = max(FRP, na.rm = TRUE),
    .groups = "drop") %>%
  mutate(density = total_hotspots / area_sqkm) %>%
  arrange(Year, desc(total_hotspots))

#change the province column name
#prov_fire_stats %>%
#  rename(province_name = adm1_name)

print(prov_fire_stats)
# save the processed data file to the project folder
write.csv(prov_fire_stats, 
          file = "C:/Users/marin/OneDrive - UBC/2. Coursework/2025 Term 2 - FRE 430/FRE 430/Final_project/Cambodia_Fire_Stats.csv", row.names = FALSE)


# create a summary table for tensity per year for each province
wide_density <- prov_fire_stats %>%
  select("adm1_name", "Year", "density") %>%
  pivot_wider(names_from = Year,
              values_from = density,
              names_prefix = "Density_", 
              )
print(wide_density)

# Stacked bar chart: fire density by province per year

# Filter to top 10 provinces by mean density for readability
top10_provinces <- prov_fire_stats %>%
  group_by(adm1_name) %>%
  summarize(mean_density = mean(density, na.rm = TRUE),
            .groups = "drop") %>%
  slice_max(mean_density, n = 10) %>%
  pull(adm1_name)

# Density version
p_density_bar <- prov_fire_stats %>%
  filter(adm1_name %in% top10_provinces) %>%
  ggplot(aes(
    x = reorder(adm1_name, -density),
    y = density,
    fill = factor(Year)
  )) +
  geom_col(position = "stack",
           colour = "white",
           linewidth = 0.3) +
  scale_fill_brewer(palette = "Set3") +
  scale_y_continuous(labels = comma) +
  labs(
    title    = "Fire hotspot density by province",
    subtitle = "Top 10 provinces — Jan–Apr fire season",
    x        = NULL,
    y        = "Fire density (hotspots / km²)",
    fill     = "Year"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title         = element_text(face = "bold"),
    plot.subtitle      = element_text(colour = "grey50"),
    panel.grid.major.x = element_blank(),
    axis.text.x        = element_text(angle = 45, hjust = 1),
    legend.position    = "right"
  )



# Proportional stacked bar (100%) to compare composition across years
p_density_pct <- prov_fire_stats %>%
  filter(adm1_name %in% top10_provinces) %>%
  ggplot(aes(x = factor(Year), y = density, fill = adm1_name)) +
  geom_col(position = "fill",
           colour = "white",
           linewidth = 0.3) +
  scale_fill_brewer(palette = "Set3") +
  scale_y_continuous(labels = percent) +
  labs(
    title    = "Proportional fire density by province per year",
    subtitle = "Top 10 provinces — Jan–Apr fire season",
    x        = "Year",
    y        = "Proportion of total density",
    fill     = "Province"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title         = element_text(face = "bold"),
    plot.subtitle      = element_text(colour = "grey50"),
    panel.grid.major.x = element_blank(),
    legend.position    = "right"
  )

p_count_bar <- prov_fire_stats %>%
  filter(adm1_name %in% top10_provinces) %>%
  ggplot(aes(
    x = reorder(adm1_name, -total_hotspots),
    y = total_hotspots,
    fill = factor(Year)
  )) +
  geom_col(position = "stack",
           colour = "white",
           linewidth = 0.3) +
  scale_fill_brewer(palette = "Set3") +
  scale_y_continuous(labels = comma) +
  labs(
    title    = "Fire hotspot count by province",
    subtitle = "Top 10 provinces — Jan–Apr fire season",
    x        = NULL,
    y        = "Total hotspots",
    fill     = "Year"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title         = element_text(face = "bold"),
    plot.subtitle      = element_text(colour = "grey50"),
    panel.grid.major.x = element_blank(),
    axis.text.x        = element_text(angle = 45, hjust = 1),
    legend.position    = "right"
  )

print(p_count_bar)
print(p_density_bar)
print(p_density_pct)

ggsave("C:/Users/marin/OneDrive - UBC/2. Coursework/2025 Term 2 - FRE 430/FRE 430/Final_project/Results/Provincial_fire_count_stacked.png",
       plot   = p_count_bar,
       width  = 10, height = 6, dpi = 150)

ggsave("C:/Users/marin/OneDrive - UBC/2. Coursework/2025 Term 2 - FRE 430/FRE 430/Final_project/Results/Provincial_fire_density-stacked.png", p_density_bar, 
       width = 10, height = 6, dpi = 150)

ggsave("C:/Users/marin/OneDrive - UBC/2. Coursework/2025 Term 2 - FRE 430/FRE 430/Final_project/Results/Provincial_fire_density-propor.png", p_density_pct, 
       width = 10, height = 6, dpi = 150)
```

#### Visualise the fire statistics for provinces

*note that hotspot data has not filtered out any low FRP, minimal brightness or for confidence level*

```{r}
#create bar chart for fire density 
ggplot(prov_fire_stats %>% filter(Year == 2024) %>% drop_na(), 
       aes(x = reorder(adm1_name, density), y = density)) +
  geom_col(fill = "#ca0020") +
  coord_flip() + # Makes province names much easier to read
  theme_minimal() +
  labs(title = "Fire density by province (2024)",
       subtitle = "Fires per square kilometer",
       x = "Province",
       y = "Density (Fires/km2)")

# a bar chart for trends over time for the top 5 provinces 
#filter for the top provinces with number of hotspots 
top_provinces <- prov_fire_stats %>%
  group_by(adm1_name) %>%
  summarize(total = sum(max_frp)) %>%
  top_n(5, total) %>%
  pull(adm1_name)

#plot the bar chart
ggplot(prov_fire_stats %>% 
         filter(adm1_name %in% 
                  top_provinces), 
       aes(x = adm1_name, 
           y = max_frp, 
           fill = factor(Year))) +
  geom_bar(stat = "identity", position = "dodge") +
  scale_fill_brewer(palette = "Set3") +
  theme_minimal() +
  labs(title = "Fire trends in the top five provinces in Cambodia",
       x = "Province",
       y = "Max FRP",
       fill = "Year")

# heatmap
ggplot(prov_fire_stats %>% drop_na(density), 
       aes(x = factor(Year), 
           y = adm1_name, 
           fill = density)) +
  geom_tile(color = "white") +
  scale_fill_viridis_c(option = "viridis", name = "Density") +
  theme_minimal() +
  labs(title = "Fire intensity by province (2019-2025)",
       subtitle = "Density in hotspots per square km",
       x = "Year",
       y = "Province")

```

Calculate proportion of uncertainty to see if there if low confidence hotspots may skew the data

Low (l) confidence daytime fire pixels are typically associated with areas of Sun glint and lower relative temperature anomaly (\<15 K) in the mid-infrared channel I4 (NASA)

```{r}

# Calculate the proportion of confidence levels per province
uncertainty_stats <- fires_in_prov %>%
  st_drop_geometry() %>%
  group_by(adm1_name, CONFIDENCE, Year) %>%
  summarize(count = n(), .groups = "drop_last") %>%
  mutate(percent = (count / sum(count)) * 100)

# View provinces with the highest percentage of 'low' confidence fires
uncertainty_stats %>% 
  filter(CONFIDENCE == "l") %>% 
  arrange(desc(percent))

ggplot(uncertainty_stats %>% drop_na(), 
       aes(x = factor(Year), 
           y = adm1_name, 
           fill = percent)) +
  geom_tile(color = "white") +
  scale_fill_viridis_c(option = "virdis", name = "percent") +
  theme_minimal() +
  labs(title = "Proportion of low confidence hotspots by province (2019-2025)",
       x = "Year",
       y = "Province")

```

Create a choropleth map

```{r}
# left join the original provinces (admin boundaries dataset) with the fire stats dataset 
# Use density
#remove any provinces with no fire hotspots
ch_map_data <- provinces %>%
  left_join(prov_fire_stats, by = "adm1_name") %>%
  mutate(density = if_else(is.na(density), 0, density))

#plot the map
ch_density <- ggplot(ch_map_data) +
  geom_sf(aes(fill = density), color = "white", size = 0.1) +
  scale_fill_distiller(palette = "YlOrRd",
                       trans = "reverse",
                       na.value = "grey90", 
                       name = "Fires/km²") +
  facet_wrap(~Year, ncol = 4) +
  theme_void() +
  labs(title = "Fire density in Cambodia (2019-2025)", 
       caption = "NASA FIRMS: VIIRS")

# use the max FRP 
#remove any provinces with no fire hotspots
ch_map_data_frp <- provinces %>%
  left_join(prov_fire_stats, by = "adm1_name") %>%
  mutate(max_frp = if_else(is.na(max_frp), 0, max_frp))

#plot the map
ch_frp <- ggplot(ch_map_data_frp) +
  geom_sf(aes(fill = max_frp), color = "white", size = 0.1) +
  scale_fill_distiller(palette = "YlOrRd",
                       trans = "reverse",
                       na.value = "grey90", 
                       name = "Max FRP") +
  facet_wrap(~Year, ncol = 4) +
  theme_void() +
  labs(title = "Fire intensity in Cambodia (2019-2025)", 
       caption = "NASA FIRMS: VIIRS")

print(ch_density)
print(ch_frp)
```

### Land use raster data

Note that lulc data is only for 2012-2023 and fire data is only for 2019-2025

```{r}
#create a list for where all the raster files are in my folder
raster_files_lulc <- list.files(path = "C:/Users/marin/OneDrive - UBC/2. Coursework/2025 Term 2 - FRE 430/FRE 430/Final_project/GEE_rasters/Cambodia_LULC_Project",
                                pattern = "\\.tif$",
                                full.names = TRUE)
#read them all in as a raster stack 
lulc_stack <- rast(raster_files_lulc)

#rename the files 
names(lulc_stack) <- c("LULC_2012","LULC_2013","LULC_2014","LULC_2015","LULC_2016","LULC_2017","LULC_2018", "LULC_2019", "LULC_2020", "LULC_2021", "LULC_2022", "LULC_2023")

print(lulc_stack)

# Check the Coordinate Reference System (CRS)
if (crs(lulc_stack) != crs(provinces)) {
  provinces <- st_transform(provinces, crs(lulc_stack))
}
plot(lulc_stack[[1]]) #2012
plot(st_geometry(provinces), 
     add = TRUE,          
     border = "black",    # Color of the province lines
     lwd = 1) 
plot(lulc_stack[[10]]) #2023
plot(st_geometry(communes), 
     add = TRUE,          
     border = "black",    # Color of the province lines
     lwd = 0.5)             # Line width

# Define LULC colour palette (one colour per class)
lulc_colours <- c(
  "1"  = "#1a6b8a",
  # Aquaculture   — deep blue
  "2"  = "#d4b483",
  # Barren        — tan
  "3"  = "#f5e642",
  # Cropland      — yellow
  "4"  = "#c8b400",
  # Crop plantation — dark yellow
  "5"  = "#a8d45a",
  # Deciduous forest — light green
  "6"  = "#2d6a4f",
  # Evergreen forest — dark green
  "7"  = "#52b788",
  # Flooded forest — teal green
  "8"  = "#74c69d",
  # Forest plantation — medium green
  "9"  = "#d8f3dc",
  # Grassland     — pale green
  "10" = "#1b4332",
  # Mangrove      — very dark green
  "11" = "#40916c",
  # Other forest  — mid green
  "12" = "#e9c46a",
  # Palm          — amber
  "13" = "#f4d35e",
  # Rice          — pale yellow
  "14" = "#e76f51",
  # Rubber        — coral
  "15" = "#b5c99a",
  # Shrub         — sage
  "16" = "#d62828",
  # Urban         — red
  "17" = "#48cae4",
  # Water         — light blue
  "18" = "#0077b6",
  # Wetland       — blue
  "19" = "#f77f00",
  # Sugarcane     — orange
  "20" = "#9b2226",
  # Cassava       — dark red
  "21" = "#ffffff",
  # Snow          — white
  "22" = "#adb5bd"   # Other         — grey
)

lulc_labels <- c(
  "1"  = "Aquaculture",
  "2"  = "Barren",
  "3"  = "Cropland",
  "4"  = "Crop plantation",
  "5"  = "Deciduous forest",
  "6"  = "Evergreen forest",
  "7"  = "Flooded forest",
  "8"  = "Forest plantation",
  "9"  = "Grassland",
  "10" = "Mangrove",
  "11" = "Other forest",
  "12" = "Palm",
  "13" = "Rice",
  "14" = "Rubber",
  "15" = "Shrub",
  "16" = "Urban",
  "17" = "Water",
  "18" = "Wetland",
  "19" = "Sugarcane",
  "20" = "Cassava",
  "21" = "Snow",
  "22" = "Other"
)

# Convert LULC 2023 raster to factor
lulc_2023 <- lulc_stack[["LULC_2023"]]
lulc_2023 <- as.factor(lulc_2023)

# Plot with tidyterra
p_lulc_map <- ggplot() +
  geom_spatraster(data = lulc_2023) +
  scale_fill_manual(
    values  = lulc_colours,
    labels  = lulc_labels,
    name    = "Land use",
    na.value = "transparent"
  ) +
  geom_sf(
    data    = provinces,
    fill    = NA,
    colour  = "white",
    linewidth = 0.4
  ) +
  labs(title    = "Land use and land cover in Cambodia 2023",
       subtitle = "RLCMS SERVIR-SEA (30m resolution)",
       caption  = "Source: Saah et al. (2020)") +
  theme_void(base_size = 12) +
  theme(
    plot.title      = element_text(face = "bold", size = 14),
    plot.subtitle   = element_text(colour = "grey50"),
    plot.caption    = element_text(colour = "grey60", size = 9),
    legend.position = "right",
    legend.key.size = unit(0.4, "cm"),
    legend.text     = element_text(size = 8)
  )

print(p_lulc_map)

ggsave("C:/Users/marin/OneDrive - UBC/2. Coursework/2025 Term 2 - FRE 430/FRE 430/Final_project/Results/LULC_2023_map.png",
       plot = p_lulc_map, width = 12, height = 8, dpi = 200)

```

### Analyse the fire and land use statistics (2019-2023)

using the fire point data, I assigned each fire point to the specific land use code

Visualized the total number of fires per lulc

```{r}
# filter the fire datasets to only include years 2019-2023
fires_lu_stats <- all_fires %>%
  filter(Year >= 2019 & Year <= 2023,
         CONFIDENCE >= 80) 
# filter the lulc stack
overlap_years <- c("LULC_2019", "LULC_2020", "LULC_2021", "LULC_2022", "LULC_2023")
lulc_subset <- lulc_stack[[overlap_years]]

# check that CRS is the same
fires_lu_stats <- st_transform(fires_lu_stats, crs(lulc_subset))

# extract values from all layers
extract_values <- terra::extract(lulc_subset, vect(fires_lu_stats))

# create a new variable for land use; using a loop to assign each fire for each year what type of land use it occured on
fires_lu_stats$land_use_code <- mapply(
 function(i, yr) {
    col_name <- paste0("LULC_", min(yr, 2023))
    extract_values[i, col_name]
  },
  i = 1:nrow(fires_lu_stats),
  yr = fires_lu_stats$Year
)

# associate the code with the lulc
fires_lu_stats <- fires_lu_stats %>%
  mutate(land_use_name = case_when(
    land_use_code == 1 ~ "Aquaculture",
    land_use_code == 2 ~ "Barren",
    land_use_code == 3 ~ "Cropland",
    land_use_code == 4 ~ "Crop Plantation",
    land_use_code == 5 ~ "Deciduous Forest",
    land_use_code == 6 ~ "Evergreen Forest",
    land_use_code == 7 ~ "Flooded Forest",
    land_use_code == 8 ~ "Forest Plantation",
    land_use_code == 9 ~ "Grassland",
    land_use_code == 10 ~ "Mangrove",
    land_use_code == 11 ~ "Other Forest",
    land_use_code == 12 ~ "Palm",
    land_use_code == 13 ~ "Rice",
    land_use_code == 14 ~ "Rubber",
    land_use_code == 15 ~ "Shrub",
    land_use_code == 16 ~ "Urban",
    land_use_code == 17 ~ "Water",
    land_use_code == 18 ~ "Wetland",
    land_use_code == 19 ~ "Sugarcane",
    land_use_code == 20 ~ "Cassava",
    land_use_code == 21 ~ "Snow",
    land_use_code == 22 ~ "Other",
    TRUE ~ "Unknown"
  ))

head(fires_lu_stats)

# spatial join to add the province to each fire incidence
#check the crs
fires_lu_stats <- st_transform(fires_lu_stats,st_crs(provinces))

#join the fire points to the provinces location
fires_lu_stats <- st_join(fires_lu_stats, 
                          provinces %>% 
                            select(adm1_name))
#check 
head(fires_lu_stats$adm1_name)

# calculate lulc statistics with fire stats 
# specifically looking at grassland, deciduous forest, cropland

frp_comparison <- fires_lu_stats %>%
  st_drop_geometry() %>%
  filter(land_use_name %in% c("Grassland", "Deciduous Forest", "Cropland")) %>%
  group_by(land_use_name) %>%
  summarize(
      avg_frp = mean(FRP, na.rm = TRUE),
      median_frp = median(FRP, na.rm = TRUE),
      sd_frp = sd(FRP, na.rm = TRUE),
      max_frp = max(FRP, na.rm = TRUE),
      total_fires = n()
    )
  
print(frp_comparison)
# find any fires that are outside of specific provincial boundaries
sum(is.na(fires_lu_stats$adm1_name))
# given that there are fires that were unmatched, use nearest neighbor to assign the unmatched to neighboring provinces

library(nngeo)  # install if needed: install.packages("nngeo")

matched    <- fires_lu_stats %>% filter(!is.na(adm1_name))
unmatched  <- fires_lu_stats %>% filter(is.na(adm1_name)) %>% select(-adm1_name)

# Find nearest province for each unmatched point
nearest_idx <- st_nn(unmatched, provinces, k = 1, returnDist = FALSE)

unmatched$adm1_name <- provinces$adm1_name[unlist(nearest_idx)]

# Recombine
fires_lu_stats <- bind_rows(matched, unmatched)

# Verify
sum(is.na(fires_lu_stats$adm1_name))  # should be 0

# Calculate the top 5 provinces  with highest fire count per land use type
top5_per_landuse <- fires_lu_stats %>%
  st_drop_geometry() %>%    
  group_by(land_use_name, adm1_name) %>%
  summarize(
    total_fires = n(),
    avg_frp     = mean(FRP, na.rm = TRUE),
    .groups     = "drop"
  ) %>%
  group_by(land_use_name) %>%
  slice_max(total_fires, n = 5) %>% # correctly picks top 5 by fire count
  ungroup() %>%
  arrange(land_use_name, desc(total_fires))

print(top5_per_landuse, 10)


# Overall summary by land use type (collapsed across provinces)
land_use_summary <- fires_lu_stats %>%
  st_drop_geometry() %>%
  filter(land_use_name %in% unique(top5_per_landuse$land_use_name)) %>% 
  group_by(land_use_name) %>%
  summarize(
    total_fires = n(),
    avg_frp     = mean(FRP, na.rm = TRUE),
    median_frp  = median(FRP, na.rm = TRUE),
    sd_frp      = sd(FRP, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  arrange(desc(total_fires))

print(land_use_summary)
p_landuse_summary <- ggplot(land_use_summary, 
                            aes(x = reorder(land_use_name, total_fires), 
                                y = total_fires, 
                                fill = land_use_name)) +
  geom_col(show.legend = FALSE, alpha = 0.9) +
  geom_text(aes(label = comma(total_fires)), 
            hjust = -0.1, size = 3.5, fontface = "bold") +
  scale_y_continuous(labels = comma, 
                     expand = expansion(mult = c(0, 0.15))) +  # extra space for labels
  scale_fill_viridis_d(option = "inferno", begin = 0.1, end = 0.9) +
  coord_flip() +
  labs(
    title    = "Total Fire Hotspots by Land Use Type",
    subtitle = "Cambodia 2019–2023",
    x        = NULL,
    y        = "Number of Hotspots"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    panel.grid.major.y = element_blank(),
    plot.title         = element_text(face = "bold"),
    plot.subtitle      = element_text(colour = "grey50")
  )

print(p_landuse_summary)

## create a summary for each year and plot it in a bar chart to compare fires per landu use type 
# Loop through each year
years <- 2019:2023

for (yr in years) {
  # Summarize for this year
  land_use_summary_yr <- fires_lu_stats %>%
    st_drop_geometry() %>%
    filter(Year == yr,
           land_use_name %in% unique(top5_per_landuse$land_use_name)) %>%
    group_by(land_use_name) %>%
    summarize(
      total_fires = n(),
      avg_frp     = mean(FRP, na.rm = TRUE),
      median_frp  = median(FRP, na.rm = TRUE),
      sd_frp      = sd(FRP, na.rm = TRUE),
      .groups     = "drop"
    ) %>%
    arrange(desc(total_fires))
  
  # Build chart
  p <- ggplot(land_use_summary_yr,
              aes(x = reorder(land_use_name, total_fires),
                  y = total_fires,
                  fill = land_use_name)) +
    geom_col(show.legend = FALSE, alpha = 0.9) +
    geom_text(aes(label = comma(total_fires)),
              hjust = -0.1, size = 3.5, fontface = "bold") +
    scale_y_continuous(labels = comma,
                       expand = expansion(mult = c(0, 0.15))) +
    scale_fill_viridis_d(option = "inferno", begin = 0.1, end = 0.9) +
    coord_flip() +
    labs(
      title    = "Total Fire Hotspots by Land Use Type (Count)",
      subtitle = paste("Cambodia", yr),
      x        = NULL,
      y        = "Number of Hotspots"
    ) +
    theme_minimal(base_size = 13) +
    theme(
      panel.grid.major.y = element_blank(),
      plot.title         = element_text(face = "bold"),
      plot.subtitle      = element_text(colour = "grey50")
    )
  print(p)
  ggsave(paste0("landuse_summary_total", yr, ".png"), p,
     width = 10, height = 6, dpi = 150)
}

# Save each chart with year in filename
# ggsave(paste0("landuse_summary_", yr, ".png"), p,
#         width = 10, height = 6, dpi = 150)

## create a summary for each year and plot it in a bar chart to compare frp per landu use type 
# Loop through each year
years <- 2019:2023

for (yr in years) {
  # Summarize for this year
  land_use_summary_yr <- fires_lu_stats %>%
    st_drop_geometry() %>%
    filter(Year == yr,
           land_use_name %in% unique(top5_per_landuse$land_use_name)) %>%
    group_by(land_use_name) %>%
    summarize(
      total_fires = n(),
      max_frp     = max(FRP, na.rm = TRUE),
      avg_frp     = mean(FRP, na.rm = TRUE),
      median_frp  = median(FRP, na.rm = TRUE),
      sd_frp      = sd(FRP, na.rm = TRUE),
      .groups     = "drop"
    ) %>%
    arrange(desc(max_frp))
  
  # chart
  p <- ggplot(land_use_summary_yr,
              aes(x = reorder(land_use_name, max_frp),
                  y = max_frp,
                  fill = land_use_name)) +
    geom_col(show.legend = FALSE, alpha = 0.9) +
    geom_text(aes(label = comma(max_frp)),
              hjust = -0.1, size = 3.5, fontface = "bold") +
    scale_y_continuous(labels = comma,
                       expand = expansion(mult = c(0, 0.15))) +
    scale_fill_viridis_d(option = "inferno", begin = 0.1, end = 0.9) +
    coord_flip() +
    labs(
      title    = "Fire Intensity by Land Use Type",
      subtitle = paste("Cambodia", yr),
      x        = NULL,
      y        = "Max FRP (MW)"
    ) +
    theme_minimal(base_size = 13) +
    theme(
      panel.grid.major.y = element_blank(),
      plot.title         = element_text(face = "bold"),
      plot.subtitle      = element_text(colour = "grey50")
    )
  print(p)
  ggsave(paste0("landuse_summary_max", yr, ".png"), p,
     width = 10, height = 6, dpi = 150)
}


# calculate the percentage change for each province year to year per land use
pct_change_table <- fires_lu_stats %>%
  st_drop_geometry() %>%
  filter(land_use_name %in% unique(top5_per_landuse$land_use_name)) %>%
  group_by(land_use_name, Year) %>%
  summarize(
    total_fires = n(),
    max_frp = max(FRP, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  arrange(land_use_name, Year) %>%
  group_by(land_use_name) %>%
  mutate(
    # Percent change from previous year
    pct_change_count = (total_fires - lag(total_fires)) / lag(total_fires) * 100,
    pct_change_frp   = (max_frp - lag(max_frp)) / lag(max_frp) * 100
  ) %>%
  ungroup() %>%
  mutate(across(c(pct_change_count, pct_change_frp), ~ round(., 1)))

# Print
print(pct_change_table)

# Formatted table
pct_change_table %>%
  mutate(
    pct_change_count = ifelse(is.na(pct_change_count), "—", paste0(pct_change_count, "%")),
    pct_change_frp   = ifelse(is.na(pct_change_frp), "—", paste0(pct_change_frp, "%"))
  ) %>%
  kable(
    col.names = c(
      "Land use",
      "Year",
      "Total fires",
      "Max FRP (MW)",
      "% change (count)",
      "% change (FRP)"
    ),
    caption   = "Year-on-year percent change in fire count and intensity by land use"
  ) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed")) %>%
  collapse_rows(columns = 1)


# Calculate percent change year to year for top 5 land uses by fire count per year
pct_change_bar <- fires_lu_stats %>%
  st_drop_geometry() %>%
  group_by(land_use_name, Year) %>%
  summarize(total_fires = n(), max_frp = max(FRP, na.rm = TRUE), .groups = "drop") %>%
  # Get top 5 land uses per year by fire count
  group_by(Year) %>%
  slice_max(total_fires, n = 5) %>%
  ungroup() %>%
  arrange(land_use_name, Year) %>%
  group_by(land_use_name) %>%
  mutate(
    pct_change_count = (total_fires - lag(total_fires)) / lag(total_fires) * 100,
    pct_change_frp   = (max_frp - lag(max_frp)) / lag(max_frp) * 100
  ) %>%
  ungroup() %>%
  filter(!is.na(pct_change_count))

# Bar chart: percent change in fire count
p_pct_count <- ggplot(pct_change_bar,
                      aes(x = reorder(land_use_name, pct_change_count),
                          y = pct_change_count,
                          fill = pct_change_count > 0)) +
  geom_col(alpha = 0.9) +
  geom_text(aes(label = paste0(round(pct_change_count, 1), "%")),
            hjust = ifelse(pct_change_bar$pct_change_count > 0, -0.1, 1.1),
            size = 3) +
  scale_fill_manual(values = c("TRUE" = "#d73027", "FALSE" = "#4575b4"),
                    labels = c("TRUE" = "Increase", "FALSE" = "Decrease"),
                    name   = NULL) +
  scale_y_continuous(expand = expansion(mult = c(0.15, 0.15))) +
  coord_flip() +
  facet_wrap(~ Year, ncol = 2) +
  labs(title    = "Year-on-year % change in fire count by land use",
       subtitle = "Top 5 land uses per year — Cambodia 2019–2023",
       x = NULL, y = "% change") +
  theme_minimal(base_size = 12) +
  theme(plot.title  = element_text(face = "bold"),
        strip.text  = element_text(face = "bold"),
        legend.position = "bottom")

# Bar chart: percent change in max FRP
p_pct_frp <- ggplot(pct_change_bar,
                    aes(x = reorder(land_use_name, pct_change_frp),
                        y = pct_change_frp,
                        fill = pct_change_frp > 0)) +
  geom_col(alpha = 0.9) +
  geom_text(aes(label = paste0(round(pct_change_frp, 1), "%")),
            hjust = ifelse(pct_change_bar$pct_change_frp > 0, -0.1, 1.1),
            size = 3) +
  scale_fill_manual(values = c("TRUE" = "#d73027", "FALSE" = "#4575b4"),
                    labels = c("TRUE" = "Increase", "FALSE" = "Decrease"),
                    name   = NULL) +
  scale_y_continuous(expand = expansion(mult = c(0.15, 0.15))) +
  coord_flip() +
  facet_wrap(~ Year, ncol = 2) +
  labs(title    = "Year-on-year % change in max FRP by land use",
       subtitle = "Top 5 land uses per year — Cambodia 2019–2023",
       x = NULL, y = "% change") +
  theme_minimal(base_size = 12) +
  theme(plot.title  = element_text(face = "bold"),
        strip.text  = element_text(face = "bold"),
        legend.position = "bottom")

print(p_pct_count)
print(p_pct_frp)

ggsave("C:/Users/marin/OneDrive - UBC/2. Coursework/2025 Term 2 - FRE 430/FRE 430/Final_project/Results/per_change_count.png", p_pct_count, 
       width = 10, height = 6, dpi = 150)

ggsave("C:/Users/marin/OneDrive - UBC/2. Coursework/2025 Term 2 - FRE 430/FRE 430/Final_project/Results/per_change_frp.png", p_pct_frp, 
       width = 10, height = 6, dpi = 150)
```

#### Temporal fire season shift via fire points

-   reindex the year to start from july to june so that the fire season can be considered starting in Dec or earlier

-   only look at the top 3 provinces with mean annual max FRP

-   need to refine the calculation for fire season

```{r}
# calculate the top 3 provinces with max frp across the 7 years
# top3_provinces <- fires_lu_stats %>%
#   st_drop_geometry() %>%
#   group_by(adm1_name) %>%
#   summarize(max_frp = max(FRP, na.rm = TRUE), .groups = "drop") %>%
#   slice_max(max_frp, n = 3) %>%
#   pull(adm1_name)
# 
# print(top3_provinces)

#calculate the top 6 provinces with mean of annual max FRP
top3_provinces <- fires_lu_stats %>%
  st_drop_geometry() %>%
  mutate(
    date      = as.Date(ACQ_DATE),
    month     = month(date),
    fire_year = ifelse(month >= 7, Year + 1, Year)
  ) %>%
  group_by(adm1_name, fire_year) %>%
  summarize(annual_max_frp = max(FRP, na.rm = TRUE),
            .groups = "drop") %>%
  group_by(adm1_name) %>%
  summarize(mean_annual_max = mean(annual_max_frp, na.rm = TRUE),
            .groups = "drop") %>%
  slice_max(mean_annual_max, n = 3) %>%
  pull(adm1_name)

print(top3_provinces)

# Aggregate max FRP by province, year, month
monthly_province <- fires_lu_stats %>%
  st_drop_geometry() %>%
  filter(adm1_name %in% top3_provinces) %>%
  mutate(
    date      = as.Date(ACQ_DATE),
    month     = month(date),
    fire_year = ifelse(month >= 7, Year + 1, Year)
  ) %>%
  group_by(adm1_name, fire_year, month) %>%
  summarize(max_frp = max(FRP, na.rm = TRUE), .groups = "drop")

# Aggregate max FRP by province, year, month, day
daily_province <- fires_lu_stats %>%
  st_drop_geometry() %>%
  filter(adm1_name %in% top3_provinces) %>%
  mutate(
    date      = as.Date(ACQ_DATE),
    doy       = yday(date),
    month     = month(date),
    fire_year = ifelse(month >= 7, Year + 1, Year)
  ) %>%
  group_by(adm1_name, fire_year, doy) %>%
  summarize(max_frp = max(FRP, na.rm = TRUE), .groups = "drop")

# Line chart — monthly FRP per year per province
p_monthly_top3 <- ggplot(monthly_province, aes(
  x = month,
  y = max_frp,
  colour = factor(fire_year)
)) +
  geom_line(linewidth = 0.9) +
  geom_point(size = 2) +
  scale_x_continuous(breaks = 1:12, labels = month.abb) +
  scale_colour_viridis_d(option = "inferno",
                         begin = 0.1,
                         end = 0.9) +
  facet_wrap( ~ adm1_name, ncol = 1, scales = "free_y") +
  labs(
    title    = "Monthly fire intensity — top 3 provinces",
    x        = NULL,
    y = "Max FRP (MW)",
    colour = "fire_year"
  ) +
  theme_minimal(base_size = 12) +
  theme(plot.title = element_text(face = "bold"),
        strip.text = element_text(face = "bold"))

# Line chart — daily FRP per year per province
p_daily_top3 <- ggplot(daily_province, aes(
  x = doy,
  y = max_frp,
  colour = factor(fire_year)
)) +
  geom_line(linewidth = 0.6, alpha = 0.8) +
  scale_x_continuous(
    breaks = c(1, 32, 60, 91, 121, 152, 182, 213, 244, 274, 305, 335),
    labels = month.abb
  ) +
  scale_colour_viridis_d(option = "inferno",
                         begin = 0.1,
                         end = 0.9) +
  facet_wrap( ~ adm1_name, ncol = 2, scales = "free_y") +
  labs(
    title = "Daily fire intensity — top 3 provinces",
    x = NULL,
    y = "Max FRP (MW)",
    colour = "fire_year"
  ) +
  theme_minimal(base_size = 12) +
  theme(plot.title = element_text(face = "bold"),
        strip.text = element_text(face = "bold"))

print(p_monthly_top3)
print(p_daily_top3)

# Function to calculate fire season breakpoints using Zhang et al. (2014) method
# Inputs:
#   daily_frp_vec: vector of daily max FRP values (ordered by day)
#   days: corresponding vector of day-of-year values
# Returns: list of k_start (beta=0.1), k_mid (beta=0.5), k_end (beta=0.9)
fire_season_breakpoints <- function(daily_frp_vec, days) {
  
  # Step 1: Calculate cumulative sum of FRP across the year
  cum_frp <- cumsum(daily_frp_vec)
  
  # Step 2: Normalise by total annual FRP (y_n in Zhang et al.)
  # This gives proportion of total fire energy released up to each day
  norm_frp <- cum_frp / sum(daily_frp_vec, na.rm = TRUE)
  
  # Step 3: Find k_beta — first day where normalised cumulative FRP exceeds threshold
  # beta = 0.1 -> season start (10% of total FRP released)
  # beta = 0.5 -> season midpoint (50% of total FRP released)
  # beta = 0.9 -> season end (90% of total FRP released)
  k_start <- days[min(which((norm_frp - 0.1) > 0))]
  k_mid   <- days[min(which((norm_frp - 0.5) > 0))]
  k_end   <- days[min(which((norm_frp - 0.9) > 0))]
  
  return(list(k_start = k_start, k_mid = k_mid, k_end = k_end))
}

# Apply breakpoint function to each province-year combination
# daily_province must be sorted by doy within each group
season_breakpoints <- daily_province %>%
  group_by(adm1_name, fire_year) %>%
  arrange(doy) %>%
  summarize(
    # Extract each breakpoint from the function output
    k_start = fire_season_breakpoints(max_frp, doy)$k_start,
    k_mid   = fire_season_breakpoints(max_frp, doy)$k_mid,
    k_end   = fire_season_breakpoints(max_frp, doy)$k_end,
    .groups = "drop"
  )

print(season_breakpoints)

# Reshape to long format for plotting all three metrics together
season_long <- season_breakpoints %>%
  pivot_longer(cols      = c(k_start, k_mid, k_end),
               names_to  = "metric",
               values_to = "doy") %>%
  # Rename for readable legend labels
  mutate(metric = recode(metric,
                         k_start = "Start (β=0.1)",
                         k_mid   = "Midpoint (β=0.5)",
                         k_end   = "End (β=0.9)"))

# Plot: track how start, midpoint, and end of fire season shift year to year
p_breakpoints <- ggplot(season_long,
                        aes(x = fire_year, y = doy, colour = metric)) +
  geom_line(linewidth = 0.9) +
  geom_point(size = 2.5) +
  # Convert doy y-axis to month labels for readability
  scale_y_continuous(breaks = c(1, 32, 60, 91, 121, 152, 182, 213, 244, 274, 305, 335),
                     labels = month.abb) +
  scale_x_continuous(breaks = 2019:2025) +
  scale_colour_manual(values = c("#f4a261", "#e76f51", "#264653")) +
  # Facet by province to compare top 6 side by side
  facet_wrap(~ adm1_name, ncol = 2) +
  labs(title    = "Fire season timing",
       subtitle = "Start (β=0.1), midpoint (β=0.5), end (β=0.9) of cumulative FRP",
       x        = "Year",
       y        = "Day of year",
       colour   = NULL) +
  theme_minimal(base_size = 12) +
  theme(plot.title = element_text(face = "bold"),
        strip.text = element_text(face = "bold"))

print(p_breakpoints)

# table for the fire season start, mid, end dates based on the cumulative daily FRP 
season_breakpoints %>%
  mutate(across(c(k_start, k_mid, k_end), 
                ~ format(as.Date(. - 1, origin = paste0(fire_year - 1, "-01-01")),   "%b %d"))) %>%
  kable(
    col.names = c("Province", "Fire year", "Start", "Midpoint", "End"),
    caption   = "Fire season breakpoints (Zhang et al. 2014)"
  ) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed")) %>%
  collapse_rows(columns = 1)


season_regression <- season_long %>%
  group_by(adm1_name, metric) %>%
  do(tidy(lm(doy ~ fire_year, data = .))) %>%
  filter(term == "fire_year") %>%
  select(adm1_name, metric, estimate, std.error, p.value) %>%
  mutate(
    direction = ifelse(estimate > 0, "Later", "Earlier"),
    sig       = case_when(
      p.value < 0.001 ~ "***",
      p.value < 0.01  ~ "**",
      p.value < 0.05  ~ "*",
      TRUE            ~ "ns"
    )
  ) %>%
  arrange(metric, p.value)

print(season_regression)
```

### Conduct pixel to pixel comparison

Rasterize by fire frequency and intensity using the count and FRP

```{r}
#create a raster template for rasterizing the fire hotspot data
# raster_template <- aggregate(lulc_stack[[1]], fact = 10, fun = "modal")
# 
# # Define where to save the outputs
output_dir <- "C:/Users/marin/OneDrive - UBC/2. Coursework/2025 Term 2 - FRE 430/FRE 430/Final_project/Fire_Rasters/"
# 
# if(!dir.exists(output_dir)) dir.create(output_dir)

# # Start the Loop
# years_to_process <- 2019:2025
# 
# for (yr in years_to_process) {
#   message(paste("Processing year:", yr))
#  annual_fires <- all_fires %>% filter(Year == yr)
#   if(nrow(annual_fires) == 0) next
#   v_fires <- vect(annual_fires) 
#  # save the frequency raster
#   rasterize(v_fires, raster_template, fun = "count", 
#             filename = paste0(output_dir, 
#                               "Fire_Count_", 
#                               yr, 
#                               ".tif"), overwrite = TRUE)
#   
#   # save the Intensity (Mean FRP) Raster
#   rasterize(v_fires, raster_template, field = "FRP", fun = "mean", 
#             filename = paste0(output_dir, 
#                               "Fire_FRP_", 
#                               yr, 
#                               ".tif"), overwrite = TRUE)
#   
#   rm(annual_fires, v_fires); gc() 
# }
# 
# message("Finished! All years saved to your folder.")
# 
# List only the count files
count_files <- list.files(output_dir,
                          pattern = "Fire_Count_.*\\.tif$",
                          full.names = TRUE)

# Create the stack
fire_count_stack <- rast(count_files)

# List only the FRP files
frp_files <- list.files(output_dir,
                        pattern = "Fire_FRP_.*\\.tif$",
                        full.names = TRUE)

# Create the stack
fire_frp_stack <- rast(frp_files)

# # Check dimensions and layer counts
# print(fire_count_stack)
# print(fire_frp_stack)

names(fire_count_stack) <- paste0("fire_count_", 2019:2025)
names(fire_frp_stack)   <- paste0("fire_frp_",   2019:2025)

# names(fire_count_stack)
# nlyr(fire_count_stack)

# check alignment for crs and resolution
crs(fire_count_stack)
crs(lulc_stack)
res(fire_count_stack)   # your fire raster: 300m
res(lulc_stack)         # LULC: ~30m

# becuase the LULC rasters are 30m 
# Use method="near" for categorical LULC data
lulc_300m <- resample(lulc_stack, fire_count_stack, method = "near")

# Verify alignment
compareGeom(fire_count_stack, lulc_300m)  # should return TRUE

# Zonal statistics — fire count and FRP per LULC type per year:
zonal_results <- list()

for (yr in 2019:2023) {
  lulc_year     <- ifelse(yr > 2023, 2023, yr)
  lulc_yr       <- lulc_300m[[paste0("LULC_", yr)]]
  fire_count_yr <- fire_count_stack[[paste0("fire_count_", yr)]]
  fire_frp_yr   <- fire_frp_stack[[paste0("fire_frp_", yr)]]

  # Zonal stats for count and FRP
  z_count <- zonal(fire_count_yr, lulc_yr, fun = "sum",  na.rm = TRUE) 
  # to calculate the annual average
  z_frp   <- zonal(fire_frp_yr, lulc_yr, fun = "max", na.rm = TRUE)

  zonal_results[[as.character(yr)]] <- tibble(
    land_use_code = z_count[[1]],
    total_fires   = z_count[[2]],
    max_frp      = z_frp[[2]],
    Year          = yr
  )
}

# Combine and add land use names:
zonal_df <- bind_rows(zonal_results) %>%
  mutate(land_use_name = case_when(
    land_use_code == 1  ~ "Aquaculture",
    land_use_code == 2  ~ "Barren",
    land_use_code == 3  ~ "Cropland",
    land_use_code == 4  ~ "Crop Plantation",
    land_use_code == 5  ~ "Deciduous Forest",
    land_use_code == 6  ~ "Evergreen Forest",
    land_use_code == 7  ~ "Flooded Forest",
    land_use_code == 8  ~ "Forest Plantation",
    land_use_code == 9  ~ "Grassland",
    land_use_code == 10 ~ "Mangrove",
    land_use_code == 11 ~ "Other Forest",
    land_use_code == 12 ~ "Palm",
    land_use_code == 13 ~ "Rice",
    land_use_code == 14 ~ "Rubber",
    land_use_code == 15 ~ "Shrub",
    land_use_code == 16 ~ "Urban",
    land_use_code == 17 ~ "Water",
    land_use_code == 18 ~ "Wetland",
    land_use_code == 19 ~ "Sugarcane",
    land_use_code == 20 ~ "Cassava",
    TRUE ~ "Unknown"
  ))

print(zonal_df)

# Define target land uses as top 5-6 by avg frp
target_land_uses_frp <- zonal_df %>%
  group_by(land_use_name) %>%
  summarize(max_frp = mean(max_frp, na.rm = TRUE), .groups = "drop") %>%
  slice_max(max_frp, n = 6) %>%
  pull(land_use_name)

# Define target land uses as top 5-6 by avg fire count
target_land_uses_count <- zonal_df %>%
  group_by(land_use_name) %>%
  summarize(total_fires = mean(total_fires, na.rm = TRUE), .groups = "drop") %>%
  slice_max(total_fires, n = 6) %>%
  pull(land_use_name)

print(target_land_uses_count)
print(target_land_uses_frp)

# visualize zonal df with top land uses with highest mean FRP
heatmap_data_frp <- zonal_df %>%
  filter(land_use_name %in% target_land_uses_frp)

# visualize zonal df with top land uses with avg annual count
heatmap_data_count <- zonal_df %>% 
  filter(land_use_name %in% target_land_uses_count)

# heat nap for avg count 
p_heat_count <- ggplot(heatmap_data_count,
                       aes(x = factor(Year),
                           y = reorder(land_use_name, 
                                       total_fires),
                           fill = total_fires)) +
  geom_tile(colour = "white", 
            linewidth = 0.5) +
  geom_text(aes(label = round(total_fires, 1)), 
            size = 3, 
            colour = "white", 
            fontface = "bold") +
  scale_fill_viridis_c(option = "cividis", 
                       labels = comma) +
  labs(title = "Total fire count by land use", 
       x = "Year", 
       y = NULL, 
       fill = "Total fires") +
  theme_minimal(base_size = 12) +
  theme(panel.grid = element_blank(), 
        plot.title = element_text(face = "bold"))

# Heatmap: avg FRP
p_heat_frp <- ggplot(heatmap_data_frp,
                     aes(x = factor(Year),
                         y = reorder(land_use_name, 
                                     max_frp),
                         fill = max_frp)) +
  geom_tile(colour = "white", linewidth = 0.5) +
  geom_text(aes(label = round(max_frp, 1)), 
            size = 3, 
            colour = "white", 
            fontface = "bold") +
  scale_fill_viridis_c(option = "cividis") +
  labs(title = "Max FRP by land use", 
       x = "Year", 
       y = NULL, 
       fill = "FRP (MW)") +
  theme_minimal(base_size = 12) +
  theme(panel.grid = element_blank(), 
        plot.title = element_text(face = "bold"))

# Bar chart: avg FRP collapsed across years 
p_frp_bar <- zonal_df %>%
  group_by(land_use_name) %>%
  summarize(max_frp = mean(max_frp, na.rm = TRUE), .groups = "drop") %>%
  filter(land_use_name %in% target_land_uses_frp) %>%
  ggplot(aes(
    x = reorder(land_use_name, max_frp),
    y = max_frp,
    fill = max_frp
  )) +
  geom_col(show.legend = FALSE, alpha = 0.9) +
  geom_text(aes(label = round(max_frp, 1)), hjust = -0.1, size = 3.5) +
  scale_fill_distiller(palette = "YlOrRd") +
  scale_y_continuous(expand = expansion(mult = c(0, 0.15))) +
  coord_flip() +
  labs(title = "Max FRP by land use (2019–2025)", x = NULL, y = "Max FRP (MW)") +
  theme_minimal(base_size = 12) +
  theme(panel.grid.major.y = element_blank(),
        plot.title = element_text(face = "bold"))

# Line chart: FRP trends over years
p_line_frp <- ggplot(heatmap_data_frp,
                     aes(x = Year, y = max_frp, colour = land_use_name)) +
  geom_line(linewidth = 0.9) +
  geom_point(size = 2) +
  scale_fill_viridis_c(option = "cividis") +
  scale_x_continuous(breaks = 2019:2025) +
  labs(
    title = "FRP trends by land use (2019–2025)",
    x = "Year",
    y = "Max FRP (MW)",
    colour = "Land use"
  ) +
  theme_minimal(base_size = 12) +
  theme(plot.title = element_text(face = "bold"))

# Line chart: fire count trends over years
p_line_count <- ggplot(heatmap_data_count,
                       aes(x = Year, y = total_fires, colour = land_use_name)) +
  geom_line(linewidth = 0.9) +
  geom_point(size = 2) +
 scale_fill_viridis_c(option = "cividis") +
  scale_x_continuous(breaks = 2019:2025) +
  labs(
    title = "Fire count trends by land use (2019–2025)",
    x = "Year",
    y = "Total fire count",
    colour = "Land use"
  ) +
  theme_minimal(base_size = 12) +
  theme(plot.title = element_text(face = "bold"))


print(p_heat_count)
print(p_heat_frp)
print(p_frp_bar)
print(p_line_frp)
print(p_line_count)

ggsave("p_heat_count.png", p_heat_count, 
       width = 14, height = 8, dpi = 150,
       path = "C:/Users/marin/OneDrive - UBC/2. Coursework/2025 Term 2 - FRE 430/FRE 430/Final_project/Results")

ggsave("p_heat_frp.png", p_heat_frp, 
       width = 14, height = 8, dpi = 150,
       path = "C:/Users/marin/OneDrive - UBC/2. Coursework/2025 Term 2 - FRE 430/FRE 430/Final_project/Results")

```

#### Regression analysis

```{r}

# Relevel land use with reference category
zonal_df <- zonal_df %>%
  mutate(land_use_name = relevel(factor(land_use_name), ref = "Evergreen Forest"))


# Ordinary least squares: fire count and land use + year
model_count <- lm(
  total_fires ~ land_use_name + Year,
  data = zonal_df %>% filter(land_use_name %in% target_land_uses_count)
)


# ordinary least squares: max frp and land use + year
model_frp <- lm(
  max_frp ~ land_use_name + Year,
  data = zonal_df %>% filter(land_use_name %in% target_land_uses_frp)
)

# summary(model_count)
# summary(model_frp)

# Stargazer table
stargazer(
  model_count,
  model_frp,
  type             = "text",
  column.labels    = c("Fire count", "Max FRP"),
  title            = "OLS regression — fire activity by land use controlling for year",
  dep.var.labels   = c("Total fires", "Max FRP (MW)"),
  omit             = "Year",
  # hide year coefficient for cleaner table
  omit.labels      = "Year FE",
  # label the omitted variable
  add.lines        = list(c("Year control", "Yes", "Yes"))
)


# OLS with interaction: fire activity ~ land use * year
model_count_int <- lm(
  total_fires ~ land_use_name * Year,
  data = zonal_df %>% filter(land_use_name %in% target_land_uses_count)
)

model_frp_int <- lm(max_frp ~ land_use_name * Year,
                    data = zonal_df %>% filter(land_use_name %in% target_land_uses_frp))

# summary(model_count_int)
# summary(model_frp_int)

stargazer(
  model_count,
  model_count_int,
  model_frp,
  model_frp_int,
  type          = "text",
  column.labels = c("Count", "Count × Year", "FRP", "FRP × Year"),
  title         = "OLS regression — fire activity by land use over time",
  add.lines     = list(c(
    "Year interaction", "No", "Yes", "No", "Yes"
  ))
)
```

#### Yearly raster analysis by province and year

```{r}
#aggregate raster fire data to province level 
#extract sum of fire count per province polygon
province_count <- extract(fire_count_stack, vect(provinces),
                          fun = "sum", na.rm = TRUE) %>% 
  as_tibble() %>% # create tibble
  mutate(adm1_name = provinces$adm1_name) %>%
  pivot_longer(-adm1_name, 
               names_to = "Year", 
               values_to = "total_fires") %>%
  mutate(Year = as.integer(str_extract(Year, "\\d{4}"))) #extract the year from the raster name

#extract max frp per province polygon
province_frp <- extract(fire_frp_stack, vect(provinces),
                        fun = "max", na.rm = TRUE) %>%
  as_tibble() %>%
  mutate(adm1_name = provinces$adm1_name) %>%
  pivot_longer(-adm1_name, 
               names_to = "Year", 
               values_to = "max_frp") %>%
  mutate(Year = as.integer(str_extract(Year, "\\d{4}")))

#join with province polygons
province_map <- provinces %>%
  left_join(province_count, by = "adm1_name") %>%
  left_join(province_frp,   by = c("adm1_name", "Year")) %>% 
  filter(!is.na(total_fires), !is.na(max_frp))

#create new tibble with total fires and max frp grouped by provinces
province_map_total <- province_map %>%
  group_by(adm1_name) %>%
  summarize(
    total_fires = sum(total_fires, na.rm = TRUE),
    max_frp = max(max_frp, na.rm = TRUE)) %>% 
  filter(!is.na(total_fires), !is.na(max_frp))

# Choropleth map for total fire count
p_choro_count <- ggplot(province_map_total) +
  geom_sf(aes(fill = total_fires),
          colour = "white",
          linewidth = 0.3) +
  scale_fill_distiller(
    palette = "YlOrRd",
    direction = 1,
    labels = comma,
    name = "Total fires"
  ) +
  labs(title    = "Fire hotspot count by province", subtitle = "Cambodia 2019–2025 (rasterized 300m)") +
  theme_void(base_size = 12) +
  theme(plot.title    = element_text(face = "bold"),
        plot.subtitle = element_text(colour = "grey50"))

#chloro map for max frp by province
p_choro_frp <- ggplot(province_map_total) +
  geom_sf(aes(fill = max_frp),
          colour = "white",
          linewidth = 0.3) +
  scale_fill_distiller(palette = "YlOrRd",
                       direction = 1,
                       name = "Max FRP (MW)") +
  labs(title    = "Fire intensity by province", subtitle = "Cambodia 2019–2025 (rasterized 300m)") +
  theme_void(base_size = 12) +
  theme(plot.title    = element_text(face = "bold"),
        plot.subtitle = element_text(colour = "grey50"))

# choropleth map for count by year
p_choro_c_yearly <- ggplot(province_map %>% filter(!is.na(Year))) +
  geom_sf(aes(fill = total_fires), colour = "white", linewidth = 0.2) +
  scale_fill_distiller(palette = "YlOrRd", direction = 1,
                       labels = comma, name = "Total fires") +
  facet_wrap(~ Year, ncol = 4) +
  labs(title = "Annual fire count by province in Cambodia") +
  theme_void(base_size = 11) +
  theme(plot.title = element_text(face = "bold"),
        strip.text = element_text(face = "bold"))

# choropleth map for max frp by year
p_choro_f_yearly <- ggplot(province_map %>% filter(!is.na(Year))) +
  geom_sf(aes(fill = max_frp), colour = "white", linewidth = 0.2) +
  scale_fill_distiller(palette = "YlOrRd", direction = 1,
                       labels = comma, name = "Max FRP") +
  facet_wrap(~ Year, ncol = 4) +
  labs(title = "Annual fire intensity by province in Cambodia") +
  theme_void(base_size = 10) +
  theme(plot.title = element_text(face = "bold"),
        strip.text = element_text(face = "bold"))

print(p_choro_c_yearly)
print(p_choro_f_yearly)
print(p_choro_count)
print(p_choro_frp)

ggsave("choropleth_fire_count.png", p_choro_c_yearly, 
       width = 14, height = 8, dpi = 150,
       path = "C:/Users/marin/OneDrive - UBC/2. Coursework/2025 Term 2 - FRE 430/FRE 430/Final_project/Results")

ggsave("choropleth_fire_intensity.png", p_choro_f_yearly, 
       width = 14, height = 8, dpi = 150,
       path = "C:/Users/marin/OneDrive - UBC/2. Coursework/2025 Term 2 - FRE 430/FRE 430/Final_project/Results")

ggsave("choropleth_hotspot_count_all.png", p_choro_count, 
       width = 14, height = 8, dpi = 150,
       path = "C:/Users/marin/OneDrive - UBC/2. Coursework/2025 Term 2 - FRE 430/FRE 430/Final_project/Results")

ggsave("choropleth_frp_all.png", p_choro_frp, 
       width = 14, height = 8, dpi = 150, 
       path = "C:/Users/marin/OneDrive - UBC/2. Coursework/2025 Term 2 - FRE 430/FRE 430/Final_project/Results")



```

#### Monthly raster stack for months for fire data

```{r}

# rasterize fire data for every month for each year
# years <- 2019:2025
# months <- 1:12
# 
# for (yr in years) {
#   for (mn in months) {
#     message(paste("Processing:", yr, "-", mn))
#     
#     # Filter for the specific month and year
#     monthly_data <- all_fires %>% filter(Year == yr, Month == mn)
#     
#     if(nrow(monthly_data) == 0) next
#     
#     # Rasterize Mean FRP and save directly to disk
#     # This keeps your RAM clean!
#     rasterize(vect(monthly_data), 
#               raster_template, 
#               field = "FRP", 
#               fun = "mean", 
#               filename = paste0(output_dir, "FRP_Mean_", yr, "_", mn, ".tif"),
#               overwrite = TRUE)
#     
#     rm(monthly_data); gc()
#   }
# }
# message("Finished! All years saved to your folder.")
# 
# List only the count files
frp_month_files <- list.files(output_dir,
                          pattern = "FRP_Mean_.*\\.tif$",
                          full.names = TRUE)

# Create the stack
frp_month_stack <- rast(frp_month_files)
print(frp_month_stack)

#check the dimensions of the CRS
terra::res(frp_month_stack) 
terra::res(lulc_subset) 
#note that the resolution for the fire data is about 300m 
# the LULC data is 30m 

# rename layers to YYYY_MM format
# Extract year and month from filenames
file_names <- basename(frp_month_files)
layer_names <- str_match(file_names, "FRP_Mean_(\\d{4})_(\\d+)\\.tif")

names(frp_month_stack) <- paste0(layer_names[,2], "_", 
                                  sprintf("%02d", as.integer(layer_names[,3])))

# Verify
# names(frp_month_stack)  # should be "2019_01", "2019_02" etc.

plot(frp_month_stack[[1]]) # 2019_1
plot(st_geometry(provinces), 
     add = TRUE,          
     border = "black",    # Color of the province lines
     lwd = 1) 
plot(frp_month_stack[[73]]) 
plot(st_geometry(provinces), 
     add = TRUE,          
     border = "black",    # Color of the province lines
     lwd = 1)             # Line width
```

Analysis for pixel to pixel comparison for monthly trends

Aiming to identify the starting month of each year that surpasses the threshold. The threshold is defined by when the max frp exceeds the 25th quantile.

```{r}
# Zonal stats per month per province
monthly_results <- list()
#for loop to extract max frp per polygon per month for each raster
for (i in 1:nlyr(frp_month_stack)) {
  #get layer name 
  layer_name <- names(frp_month_stack)[i]
  yr  <- as.integer(substr(layer_name, 1, 4))
  mn  <- as.integer(substr(layer_name, 6, 7))
  # extract year and month from the layer name
  extracted <- extract(frp_month_stack[[i]], vect(provinces),
                       fun = "max", na.rm = TRUE) %>%
    #extract max frp per province polygon for the month
    as_tibble() %>%
    select(1,2) %>% 
    # add province name 
    mutate(adm1_name = provinces$adm1_name,
           Year      = yr,
           Month     = mn)
  #store result in the list using layer name 
  monthly_results[[layer_name]] <- extracted
}

#combine all months and years into one dataframe
monthly_df <- bind_rows(monthly_results) 
 # %>%
 #  rename(max_frp = 2) %>% #rename column 2 to max frp 
 #  filter(!is.na(max_frp)) #remove rows with no fire activity

names(monthly_df)[1:2] <- c("ID", "max_frp")
monthly_df <- monthly_df %>% 
  select(adm1_name, Year, Month, max_frp) %>%
  filter(!is.na(max_frp))

head(monthly_df)
monthly_df %>% count(Year, Month)

# Get top 3 provinces by max FRP from raster data
top3_provinces <- monthly_df %>%
  group_by(adm1_name) %>%
  summarize(max_frp = max(max_frp, na.rm = TRUE)) %>%
  slice_max(max_frp, n = 3) %>%
  pull(adm1_name)

print(top3_provinces)

# Define fire season start per year
# define the threshold 
threshold <- quantile(monthly_df$max_frp, 0.25, na.rm = TRUE)

# Filter and plot
season_top3 <- monthly_df %>%
  filter(adm1_name %in% top3_provinces) %>% 
  group_by(adm1_name) %>%
  mutate(threshold = quantile(max_frp, 0.25, na.rm = TRUE)) %>%
  ungroup() %>%
  filter(max_frp >= threshold)

p_season_top3 <- ggplot(season_top3,
                        aes(x = Month, y = max_frp, colour = factor(Year))) +
  geom_line(linewidth = 0.9) +
  geom_point(size = 2) +
  scale_x_continuous(breaks = 1:12, labels = month.abb) +
  scale_colour_viridis_d(option = "inferno", begin = 0.1, end = 0.9) +
  facet_wrap(~ adm1_name, ncol = 1, scales = "free_y") +
  labs(title    = "Monthly fire intensity by year — top 3 provinces",
       subtitle = "Max FRP per month (rasterized 300m)",
       x = NULL, y = "Max FRP (MW)", colour = "Year") +
  theme_minimal(base_size = 12) +
  theme(plot.title = element_text(face = "bold"),
        strip.text = element_text(face = "bold"))

print(p_season_top3)
```

Add Raster for human modification and population

```{r}
#create a list for where all the raster files are in my folder
raster_files_pop <- list.files(path = "C:/Users/marin/OneDrive - UBC/2. Coursework/2025 Term 2 - FRE 430/FRE 430/Final_project/GEE_rasters/Cambodia_pop",
                                pattern = "\\.tif$",
                                full.names = TRUE)
#read them all in as a raster stack 
Pop_stack <- rast(raster_files_pop)

#rename the files 
names(Pop_stack) <- c("HM_2016","Pop_count")

# Separate HFI and population
HFI_raster <- Pop_stack[["HM_2016"]]
Pop_raster  <- Pop_stack[["Pop_count"]]

# Check
print(HFI_raster)
plot(HFI_raster)

```

#### HFI analysis with fire data

```{r}
# Reproject HFI to match fire raster CRS 
HFI_raster <- project(HFI_raster, crs(fire_frp_stack))

# Aggregate fire rasters from 300m to 1km to match HFI
fire_frp_1km   <- resample(fire_frp_stack,   HFI_raster, method = "bilinear")
fire_count_1km <- resample(fire_count_stack, HFI_raster, method = "sum")

# Verify alignment
compareGeom(fire_frp_1km, HFI_raster)

# Extract HFI and fire values per province
#fire count 
province_hfi_count <- extract(fire_count_1km, vect(provinces),
                               fun = "sum", na.rm = TRUE) %>%
  as_tibble() %>%
  mutate(adm1_name = provinces$adm1_name) %>%
  pivot_longer(-adm1_name, names_to = "Year", values_to = "total_fires") %>%
  mutate(Year = as.integer(str_extract(Year, "\\d{4}")))
#fire frp
province_hfi_frp <- extract(fire_frp_1km, vect(provinces),
                             fun = "max", na.rm = TRUE) %>%
  as_tibble() %>%
  mutate(adm1_name = provinces$adm1_name) %>%
  pivot_longer(-adm1_name, names_to = "Year", values_to = "max_frp") %>%
  mutate(Year = as.integer(str_extract(Year, "\\d{4}")))
# hfi raster
province_hfi_val <- extract(HFI_raster, vect(provinces),
                             fun = "mean", na.rm = TRUE) %>%
  as_tibble() %>%
  mutate(adm1_name = provinces$adm1_name) %>%
  rename(mean_hfi = 2)

# Join all together
province_hfi_sf <- provinces %>%
  left_join(province_hfi_count %>%
              group_by(adm1_name) %>%
              summarize(total_fires = sum(total_fires, na.rm = TRUE),
                        .groups = "drop"), by = "adm1_name") %>%
  left_join(province_hfi_frp %>%
              group_by(adm1_name) %>%
              summarize(max_frp = max(max_frp, na.rm = TRUE),
                        .groups = "drop"), by = "adm1_name") %>%
  left_join(province_hfi_val, by = "adm1_name") %>%
  filter(!is.na(mean_hfi), !is.na(max_frp))

# Spearman correlation 
cor.test(province_hfi_sf$mean_hfi, province_hfi_sf$max_frp,
         method = "spearman")
cor.test(province_hfi_sf$mean_hfi, province_hfi_sf$total_fires,
         method = "spearman")

# geographic weighted regression using GWmodel
province_sp_hfi <- as(province_hfi_sf, "Spatial")
#bandwidth selection for basic GWR; using Akaike information criterion (AIC)
#total fires ~ HFI
bw_count    <- bw.gwr(
  total_fires ~ mean_hfi,
  data = province_sp_hfi,
  approach = "AIC",
  adaptive = TRUE
)
gwr_count   <- gwr.basic(
  total_fires ~ mean_hfi,
  data = province_sp_hfi,
  bw = bw_count,
  adaptive = TRUE
)
print(gwr_count)

# max frp ~ HFI
bw_frp      <- bw.gwr(
  max_frp ~ mean_hfi,
  data = province_sp_hfi,
  approach = "AIC",
  adaptive = TRUE
)
gwr_frp     <- gwr.basic(max_frp ~ mean_hfi,
                         data = province_sp_hfi,
                         bw = bw_frp,
                         adaptive = TRUE)
print(gwr_frp)

# results
gwr_count_results <- as.data.frame(gwr_count$SDF) %>%
  mutate(adm1_name = province_hfi_sf$adm1_name) %>%
  select(adm1_name, Local_R2, mean_hfi) %>%
  rename(Local_R2_count = Local_R2, mean_hfi_coef_count = mean_hfi)

gwr_frp_results <- as.data.frame(gwr_frp$SDF) %>%
  mutate(adm1_name = province_hfi_sf$adm1_name) %>%
  select(adm1_name, Local_R2, mean_hfi) %>%
  rename(Local_R2_frp = Local_R2, mean_hfi_coef_frp = mean_hfi)

# join results to the map
province_hfi_map <- province_hfi_sf %>%
  left_join(gwr_count_results, by = "adm1_name") %>%
  left_join(gwr_frp_results, by = "adm1_name")

# HFI choropleth
p_hfi_choro <- ggplot(province_hfi_map) +
  geom_sf(aes(fill = mean_hfi),
          colour = "white",
          linewidth = 0.3) +
  scale_fill_distiller(palette = "YlOrRd",
                       direction = 1,
                       name = "Mean HFI") +
  labs(title = "Human footprint index by province", subtitle = "Cambodia (WCS/Columbia, 2016)") +
  theme_void(base_size = 12) +
  theme(plot.title = element_text(face = "bold"),
        plot.subtitle = element_text(colour = "grey50"))

# total fires choropleth
p_gwr_hfi_count <- ggplot(province_hfi_map) +
  geom_sf(aes(fill = total_fires),
          colour = "white",
          linewidth = 0.3) +
  scale_fill_distiller(
    palette = "YlOrRd",
    direction = 1,
    labels = comma,
    name = "Total fires"
  ) +
  labs(title    = "Total fire count by province", subtitle = "Cambodia 2019–2025 (1km resolution)") +
  theme_void(base_size = 12) +
  theme(plot.title = element_text(face = "bold"),
        plot.subtitle = element_text(colour = "grey50"))

# GWR Local R² — fire count
p_gwr_r2_count <- ggplot(province_hfi_map) +
  geom_sf(aes(fill = Local_R2_count),
          colour = "white",
          linewidth = 0.3) +
  scale_fill_distiller(palette = "YlOrRd",
                       direction = 1,
                       name = "Local R²") +
  labs(title    = "GWR local R² — total fires ~ HFI", subtitle = "Cambodia 2019–2025") +
  theme_void(base_size = 12) +
  theme(plot.title = element_text(face = "bold"),
        plot.subtitle = element_text(colour = "grey50"))

# GWR Local R² — max FRP
p_gwr_r2_frp <- ggplot(province_hfi_map) +
  geom_sf(aes(fill = Local_R2_frp),
          colour = "white",
          linewidth = 0.3) +
  scale_fill_distiller(palette = "YlOrRd",
                       direction = 1,
                       name = "Local R²") +
  labs(title    = "GWR local R² — max FRP ~ HFI", subtitle = "Cambodia 2019–2025") +
  theme_void(base_size = 12) +
  theme(plot.title = element_text(face = "bold"),
        plot.subtitle = element_text(colour = "grey50"))

# Print all four panels
p_hfi_choro + p_gwr_hfi_count + p_gwr_r2_count + p_gwr_r2_frp +
  plot_layout(ncol = 2) +
  plot_annotation(title = "HFI vs fire activity by province — Cambodia",
                  theme = theme(plot.title = element_text(face = "bold")))

print(p_hfi_choro)
print(p_gwr_hfi_count)
print(p_gwr_r2_count)
print(p_gwr_r2_frp)

# calculate GWR for one year to see if the resulting R2 differs significantly from the sum total across the 5 years
# Filter province_map to single year
province_2020 <- province_map %>%
  filter(Year == 2020) %>%
  left_join(province_hfi_val, by = "adm1_name") %>%
  filter(!is.na(mean_hfi), !is.na(total_fires))

# GWR 
province_sp_2020 <- as(province_2020, "Spatial")

bw_2020 <- bw.gwr(total_fires ~ mean_hfi, data = province_sp_2020,
                  approach = "AIC", adaptive = TRUE)

gwr_2020 <- gwr.basic(total_fires ~ mean_hfi, data = province_sp_2020,
                      bw = bw_2020, adaptive = TRUE)

print(gwr_2020)


```

#### GWR for HFI \~ total fires at the commune level

```{r}
# Extract total fires per commune for 2020 only
commune_count_2020 <- extract(fire_count_1km[["fire_count_2020"]], vect(communes),
                               fun = "sum", na.rm = TRUE) %>%
  as_tibble() %>%
  mutate(adm3_name = communes$adm3_name)

# xtract mean HFI per commune
commune_hfi <- extract(HFI_raster, vect(communes),
                       fun = "mean", na.rm = TRUE) %>%
  as_tibble() %>%
  mutate(adm3_name = communes$adm3_name) %>%
  rename(mean_hfi = 2)

# Join 
commune_map <- communes %>%
  left_join(commune_count_2020 %>% rename(total_fires = 2),
            by = "adm3_name") %>%
  left_join(commune_hfi, by = "adm3_name") %>%
  filter(!is.na(mean_hfi), !is.na(total_fires))

# Spearman correlation 
cor.test(commune_map$total_fires, commune_map$mean_hfi, method = "spearman")

# GWR
commune_sp <- as(commune_map, "Spatial")

bw_commune <- bw.gwr(total_fires ~ mean_hfi, data = commune_sp,
                     approach = "AIC", adaptive = TRUE)

gwr_commune <- gwr.basic(total_fires ~ mean_hfi, data = commune_sp,
                         bw = bw_commune, adaptive = TRUE)

print(gwr_commune)

# map results
gwr_commune_results <- as.data.frame(gwr_commune$SDF) %>%
  mutate(adm3_name = commune_map$adm3_name)

commune_gwr_map <- commune_map %>%
  left_join(gwr_commune_results %>%
              select(adm3_name, Local_R2, mean_hfi) %>%
              rename(mean_hfi_coef = mean_hfi),
            by = "adm3_name")

# Plots 
p_commune_hfi <- ggplot(commune_gwr_map) +
  geom_sf(aes(fill = mean_hfi), colour = NA) +
  scale_fill_distiller(palette = "YlOrRd", direction = 1, name = "Mean HFI") +
  labs(title    = "Human footprint index by commune",
       subtitle = "Cambodia (WCS/Columbia, 2016)") +
  theme_void(base_size = 12) +
  theme(plot.title = element_text(face = "bold"))

p_commune_fires <- ggplot(commune_gwr_map) +
  geom_sf(aes(fill = total_fires), colour = NA) +
  scale_fill_distiller(palette = "YlOrRd", direction = 1,
                       labels = comma, name = "Total fires") +
  labs(title    = "Total fire count by commune",
       subtitle = "Cambodia 2020") +
  theme_void(base_size = 12) +
  theme(plot.title = element_text(face = "bold"))

p_commune_r2 <- ggplot(commune_gwr_map) +
  geom_sf(aes(fill = Local_R2), colour = NA) +
  scale_fill_distiller(palette = "YlOrRd", direction = 1, name = "Local R²") +
  labs(title    = "GWR local R² — total fires ~ HFI",
       subtitle = "Cambodia commune level 2020") +
  theme_void(base_size = 12) +
  theme(plot.title = element_text(face = "bold"))



print(p_commune_hfi)
print(p_commune_fires)
print(p_commune_r2)
```
