---
title: "430_project"
output: html_document
---

## FRE 490 Spatial Data Analysis and Remote Sensing

##### Instructor: Jonathan Proctor

##### Teaching Assistant: Bhavita Ragoo

### Final Project workflow

-   This project entails: 7-minute presentation,
-   written analysis documenting the project and the results (5-pages double spaced text plus figures, brief intro/research question/methods section/results

### Objectives:

In this project, I will design a research question and then conduct the analysis and write up the results. Initial questions (based on proposal):

1)  How does air quality vary across different land use types in the lower Mekong region?

2)  Does air quality disproportionately impact vulnerable/ high population density areas?

#### *Refined question:*

*To what extent do seasonal fire hotspots (MODIS/VIIRS) correlate with specific land use types in Cambodia?*

*What are the demographic profiles of the populations most frequently exposed to these events? What is the proximity to areas with high human development index?*

side quest: What are the implications of air quality and downwind potentially vulnerable communities?

#### Hypothesis

Areas with higher fire density occurs in agricultural land use types and in transitional areas (proximate to cropland and forested areas). Increased fire activity occurs in areas with high human footprint index.

### Datasets

1)  Active fire points: NASA VIIRS (375m resolution): 2012-present; thermal anomalies [link](https://firms.modaps.eosdis.nasa.gov/map/)

    1)  FIRMS -\> GEE: <https://code.earthengine.google.com/?scriptPath=Examples%3ADatasets%2FFIRMS%2FFIRMS>

    2)  Considering using Burned area with MODIS -\> GEE: <https://code.earthengine.google.com/?scriptPath=Examples%3ADatasets%2FMODIS%2FMODIS_061_MCD64A1>

2)  Land use: Regional land cover monitoring system (RLCMS) by SERVIR-SEA (30m resolution) (2012 to 2023) [link](https://landcovermapping.org/en/about/) (Saah et al 2020)

    1)  GEE: <https://code.earthengine.google.com/ddc999619aa0c29514cccfc673a4f7e1>

3)  Cambodian spatial distribution of population and age and gender structures from WorldPop (100m) (2015-2020) [link](https://data.humdata.org/organization/worldpop?groups=khm&q=&ext_page_size=25)

    1)  Estimated population GEE: <https://code.earthengine.google.com/25af8a8783b214b89b8f5bef55e1be80>

4)  Cambodia administrative boundaries: Latest data received 2024 [link](https://data.humdata.org/dataset/cod-ab-khm)

    1)  Natural protected areas 1993-2023: [link](https://data.opendevelopmentmekong.net/en/dataset/protectedareas) national park, wildlife sanctuary, protected landscape, multiple use, natural heritage site, Ramsar site, biosphere reserve, marine national park

5)  Human footprint index: relative human influence in each terrestrial biome (%); updated to 2025; 1 km resolution created from human population density, human land use and infrastructure (built up areas, nighttime lights, land use/land cover), human access; least influence - 0%, most influenced - 100%; WCS/Columbia University [link](https://hub.arcgis.com/maps/65518e782be04e7db31de65d53d591a9/about)

    1)  GEE: (Kennedy et al., 2019) <https://code.earthengine.google.com/80e4e1006b6ec26ece0975766adc57b1>

6)  OpenAQ: ground validation for air pollution detected from local devices [link](https://explore.openaq.org/#3.63/12.44/99.08)

7)  Poverty and vulnerability map for Cambodia: AI-calculated using dimensions that include education, health, living standard, employment, monetary; launched in 2024 [link](https://cambodiapovertymapping.sig-gis.com/en/home/)

### Analysis

Fire hotspot density per commune

land cover/ use % per commune

correlation

raster::overlay with population density and human footprint index –\> regression analysis

1.  Cross sectional modeling
2.  cross validation regression
3.  moving window regression
4.  panel modeling

### Future Work

This preliminary analysis aims to incorporate on-the-ground data validation and novel insights on land use, hidden social-ecological values, and human-interaction landscape dynamics with an emphasis on fire use and livelihoods. This data collection will use the [LiteFarm](https://r9-kdd3.math.ubc.ca/home/) application.

## Setup

Install and load necessary packages:

```{r}
rm(list = ls())
library(raster) # handle raster data
library(ggplot2) #data visulations
library(ggthemes) #enhance data vis
library(stargazer) # regression outputs
library(cowplot) # putting plots together
library(sf) # handling vector operations - spatial features
library(dplyr) # data wrangling
library(scales) #graphical scales for vis
library(tidyr) # data wrangling
library(lubridate) #parsing dates
library(terra) #handle raster data (faster)

setwd("C:/Users/marin/OneDrive - UBC/2. Coursework/2025 Term 2 - FRE 430/FRE 430/Final_project")
```

## Preliminary Analysis

### Fire activity

Standardize the data sets to the same coordinate system: WGS 84 / UTM zone 48N (EPSG:32648)

Temporal window: Fire season from January 1st to April 30th for years: 2019 to 2025 using VIIRS

```{r}
#focus on a specific folder with the dataset
file_list_fires <- list.files(path = "C:/Users/marin/OneDrive - UBC/2. Coursework/2025 Term 2 - FRE 430/FRE 430/Final_project/Cambodia_dl-03.01_viirs", 
                        pattern = "\\.shp$", 
                        full.names = TRUE)
# Read in all the files at one time using lapply
lapply(file_list_fires, function(x) names(st_read(x, quiet = TRUE)))
# read the files in as sf objects
sf_list <- lapply(file_list_fires, read_sf)
# combine all the years (2018 April - 2026 Jan) into one dataframe
all_fires <- bind_rows(sf_list)


# filter the dataframe for the fire season of Jan-April
all_fires <- all_fires %>%
  mutate(ACQ_DATE = as.Date(ACQ_DATE), # Convert character/factor to Date
         Month = month(ACQ_DATE),     # Extract month as a number (1-12)
         Year = year(ACQ_DATE))  # Extract year as a number) %>%

# create variable filtered for only fire season 
all_fires_jan_apr <- all_fires %>%
  filter(Year != 2018 & Year != 2026) %>%
  filter(Month >= 1 & Month <= 4) 

fire_season_counts <- all_fires_jan_apr %>%
  mutate(Year = year(ACQ_DATE)) %>%
  group_by(Year) %>%
  summarize(season_fires = n())

head(fire_season_counts)
```

```{r}
#plot the fires for each year
ggplot() +
geom_sf(data = all_fires_jan_apr, aes(color = BRIGHT_T31), 
        size = 0.5, alpha = 0.7) +
  # grid by year
  facet_wrap(~ Year, ncol = 4) +  
  scale_color_viridis_c(option = "inferno", name = "BRIGHTNESS") +
  theme_void() +
  labs(title = "VIIRS Fire Hotspots in Cambodia (Jan-Apr)",
       caption = "Source: FIRMS VIIRS 375m") +
  coord_sf() # Ensures the map isn't stretched
```

### Admin boundaries layer: provinces and communes

```{r}

# using the latest data for administrative boundaries - there are three levels - province, district, commune

file_list_admin <- list.files(path = "C:/Users/marin/OneDrive - UBC/2. Coursework/2025 Term 2 - FRE 430/FRE 430/Final_project/Cambodia_admin", 
                        pattern = "\\.shp$", 
                        full.names = TRUE)
# Read in all the files at one time using lapply
lapply(file_list_admin, function(x) names(st_read(x, quiet = TRUE)))
# read the files in as sf objects
sf_list_admin <- lapply(file_list_admin, read_sf)
# rename each file for the specific administrative boundary
country <- sf_list_admin[[1]]
provinces <- sf_list_admin[[2]]
districts <- sf_list_admin[[3]]
communes  <- sf_list_admin[[4]]
# simple plot for the provinces
ggplot()+
  geom_sf(data = provinces, fill = "gray95", color = "black") +
  theme_void() +
  labs(title = "Provinces of Cambodia")
# plot the communes
ggplot()+
  geom_sf(data = communes, fill = "gray95", color = "black") +
  theme_void() +
  labs(title = "Communes of Cambodia")
# count communes per province
commune_per_province <- communes %>%
  st_drop_geometry() %>%
  group_by(adm1_name) %>%
  summarise(num_communes = n()) %>%
  arrange(desc(num_communes))

print(commune_per_province)


```

### Provincial analysis: fire activity

-   Hotspot count: measuring frequency

-   mean FRP: identifying high intensity fires vs. smaller fires

-   Fire density: hotspots per km2: normalizes the data across land area

```{r}
# confirm that the datasets are using the same CRS
if (st_crs(all_fires_jan_apr) != st_crs(provinces)) {
  warning("CRS mismatch detected! Reprojecting fires to match provinces...")
  all_fires_jan_apr <- st_transform(all_fires_jan_apr, st_crs(provinces))
} else {
  message("CRS is in alignment!")
}
#repair administrative boundaries to ensure no overlapping boundaries
provinces <- st_make_valid(provinces)
# spatial join to ensure that each fire point is attributed to a province where it is located
fires_in_prov <- st_join(all_fires_jan_apr, provinces)

# aggregate by year and province 
# calculate total count, mean FRP, and fire density
prov_fire_stats <- fires_in_prov %>%
  st_drop_geometry() %>%    # Dropping geometry makes the calculation 100x faster
  group_by(Year, adm1_name, area_sqkm) %>% 
  summarize(
    total_hotspots = n(),
    avg_frp = mean(FRP, na.rm = TRUE),
    max_frp = max(FRP, na.rm = TRUE),
    .groups = "drop") %>%
  mutate(density = total_hotspots / area_sqkm) %>%
  arrange(Year, desc(total_hotspots))

#change the province column name
#prov_fire_stats %>%
#  rename(province_name = adm1_name)

print(prov_fire_stats)
# save the processed data file to the project folder
write.csv(prov_fire_stats, 
          file = "C:/Users/marin/OneDrive - UBC/2. Coursework/2025 Term 2 - FRE 430/FRE 430/Final_project/Cambodia_Fire_Stats.csv", row.names = FALSE)


# create a summary table for tensity per year for each province
wide_density <- prov_fire_stats %>%
  select("adm1_name", "Year", "density") %>%
  pivot_wider(names_from = Year,
              values_from = density,
              names_prefix = "Density_", 
              )
print(wide_density)
```

Visualise the fire statistics for provinces

*note that hotspot data has not filtered out any low FRP, minimal brightness or for confidence level*

```{r}
#create bar chart for fire density 
ggplot(prov_fire_stats %>% filter(Year == 2024) %>% drop_na(), 
       aes(x = reorder(adm1_name, density), y = density)) +
  geom_col(fill = "#ca0020") +
  coord_flip() + # Makes province names much easier to read
  theme_minimal() +
  labs(title = "Fire density by province (2024)",
       subtitle = "Fires per square kilometer",
       x = "Province",
       y = "Density (Fires/km2)")

# a Line graph for trends over time for the top 5 provinces 

top_provinces <- prov_fire_stats %>%
  group_by(adm1_name) %>%
  summarize(total = sum(total_hotspots)) %>%
  top_n(5, total) %>%
  pull(adm1_name)

ggplot(prov_fire_stats %>% 
         filter(adm1_name %in% 
                  top_provinces), 
       aes(x = Year, 
           y = total_hotspots, 
           color = adm1_name)) +
  geom_line(size = 1) +
  geom_point() +
  theme_minimal() +
  labs(title = "Fire trends in the top five provinces in Cambodia",
       y = "Total Hotspot Count",
       color = "Province")

# heatmap
ggplot(prov_fire_stats %>% drop_na(density), 
       aes(x = factor(Year), 
           y = adm1_name, 
           fill = density)) +
  geom_tile(color = "white") +
  scale_fill_viridis_c(option = "virdis", name = "Density") +
  theme_minimal() +
  labs(title = "Fire intensity by province (2019-2025)",
       subtitle = "Density in hotspots per square km",
       x = "Year",
       y = "Province")

```

Calculate proportion of uncertainty to see if there if low confidence hotspots may skew the data

Low (l) confidence daytime fire pixels are typically associated with areas of Sun glint and lower relative temperature anomaly (\<15 K) in the mid-infrared channel I4 (NASA)

```{r}

# Calculate the proportion of confidence levels per province
uncertainty_stats <- fires_in_prov %>%
  st_drop_geometry() %>%
  group_by(adm1_name, CONFIDENCE, Year) %>%
  summarize(count = n(), .groups = "drop_last") %>%
  mutate(percent = (count / sum(count)) * 100)

# View provinces with the highest percentage of 'low' confidence fires
uncertainty_stats %>% 
  filter(CONFIDENCE == "l") %>% 
  arrange(desc(percent))

ggplot(uncertainty_stats %>% drop_na(), 
       aes(x = factor(Year), 
           y = adm1_name, 
           fill = percent)) +
  geom_tile(color = "white") +
  scale_fill_viridis_c(option = "virdis", name = "percent") +
  theme_minimal() +
  labs(title = "Proportion of low confidence hotspots by province (2019-2025)",
       x = "Year",
       y = "Province")

```

Create a choropleth map

```{r}
# left join the original provinces (admin boundaries dataset) with the fire stats dataset 
#remove any provinces with no fire hotspots
ch_map_data <- provinces %>%
  left_join(prov_fire_stats, by = "adm1_name") %>%
  mutate(density = if_else(is.na(density), 0, density))

#plot the map
ggplot(ch_map_data) +
  geom_sf(aes(fill = density), color = "white", size = 0.1) +
  scale_fill_viridis_c(option = "inferno", 
                       na.value = "grey90", 
                       name = "Fires/km²") +
  facet_wrap(~Year, ncol = 4) +
  theme_void() +
  labs(title = "Fire intensity in Cambodia (2019-2025)", 
       caption = "NASA FIRMS: VIIRS")

```

### Land use raster data

Note that lulc data is only for 2012-2023 and fire data is only for 2019-2025

```{r}
#create a list for where all the raster files are in my folder
raster_files_lulc <- list.files(path = "C:/Users/marin/OneDrive - UBC/2. Coursework/2025 Term 2 - FRE 430/FRE 430/Final_project/GEE_rasters/Cambodia_LULC_Project",
                                pattern = "\\.tif$",
                                full.names = TRUE)
#read them all in as a raster stack 
lulc_stack <- rast(raster_files_lulc)

#rename the files 
names(lulc_stack) <- c("LULC_2012","LULC_2013","LULC_2014","LULC_2015","LULC_2016","LULC_2017","LULC_2018", "LULC_2019", "LULC_2020", "LULC_2021", "LULC_2022", "LULC_2023")

print(lulc_stack)

# Check the Coordinate Reference System (CRS)
if (crs(lulc_stack) != crs(provinces)) {
  provinces <- st_transform(provinces, crs(lulc_stack))
}

```

### Analyse the fire and land use statistics (2019-2023)

```{r}
# filter the fire datasets to only include years 2019-2023
fires_lu_stats <- all_fires_jan_apr %>%
  filter(Year >= 2019 & Year <= 2023)
# filter the lulc stack
overlap_years <- c("LULC_2019", "LULC_2020", "LULC_2021", "LULC_2022", "LULC_2023")
lulc_subset <- lulc_stack[[overlap_years]]

# check that CRS is the same
fires_lu_stats <- st_transform(fires_lu_stats, crs(lulc_subset))

# extract values from all layers
extract_values <- terra::extract(lulc_subset, vect(fires_lu_stats))

# create a new variable for land use; using a loop to assign each fire for each year what type of land use it occured on
fires_lu_stats$land_use_code <- sapply(
  1:nrow(fires_lu_stats), 
  function(i) {
  col_name <- paste0("LULC_", 
                     fires_lu_stats$Year[i])
  return(extract_values[i, 
                        col_name])
})

# associate the code with the lulc
fires_lu_stats <- fires_lu_stats %>%
  mutate(land_use_name = case_when(
    land_use_code == 1 ~ "Aquaculture",
    land_use_code == 2 ~ "Barren",
    land_use_code == 3 ~ "Cropland",
    land_use_code == 4 ~ "Crop Plantation",
    land_use_code == 5 ~ "Deciduous Forest",
    land_use_code == 6 ~ "Evergreen Forest",
    land_use_code == 7 ~ "Flooded Forest",
    land_use_code == 8 ~ "Forest Plantation",
    land_use_code == 9 ~ "Grassland",
    land_use_code == 10 ~ "Mangrove",
    land_use_code == 11 ~ "Other Forest",
    land_use_code == 12 ~ "Palm",
    land_use_code == 13 ~ "Rice",
    land_use_code == 14 ~ "Rubber",
    land_use_code == 15 ~ "Shrub",
    land_use_code == 16 ~ "Urban",
    land_use_code == 17 ~ "Water",
    land_use_code == 18 ~ "Wetland",
    land_use_code == 19 ~ "Sugarcane",
    land_use_code == 20 ~ "Cassava",
    land_use_code == 21 ~ "Snow",
    land_use_code == 22 ~ "Other",
    TRUE ~ "Unknown"
  ))

head(fires_lu_stats)
```

Add Raster for human modification and population

```{r}
#create a list for where all the raster files are in my folder
raster_files_pop <- list.files(path = "C:/Users/marin/OneDrive - UBC/2. Coursework/2025 Term 2 - FRE 430/FRE 430/Final_project/GEE_rasters/Cambodia_pop",
                                pattern = "\\.tif$",
                                full.names = TRUE)
#read them all in as a raster stack 
Pop_stack <- rast(raster_files_pop)

#rename the files 
names(Pop_stack) <- c("HM_2016","Pop_count")

print(Pop_stack)

```
