---
title: "Progress_update"
output: 
  html_document:
    toc: true
  github_document: 
    toc: true
---

# Geospatial Analysis: Practice

**Author:** Marina

**Date:** `r Sys.Date()`

## Key Updates

-   Proposed date for spring review: 29 April morning

-   430 final project presentations: 4/7 or 4/9 (write up due 4/16)

-   Working with LiteFarm developers to translate app using CrowdIn

    -   contracting translators for Khmer

    -   sketching out potential preliminary concept for exp design

    -   ask about the typical frequency of data input \*\*

## .Rmd Overview

This .rmd contains R code and output for practicing geospatial data analysis and creating the poster for the WDCAG 2026 and trial maps from field data.

### Loading libraries

```{r}
library(sf)
library(raster); library(dplyr); library(ggplot2); library(scales)
library(terra)
# set working directory to where the fire data is saved 
setwd("C:/Users/marin/Documents/1. Liu Lab/Practice_data_folder/cambodia_field_sites")
#load in datasets using sf package
savanna_plots <- st_read("savanna_plots.gpkg")
forest_plots <- st_read("forest_plots.gpkg")
kr_2024_rgb <- rast("kr_2024_rgb.tif")

```

### Plotting the field data for field sites in Cambodia

*note that the raster should be flipped 180 degrees*

```{r}

# using base terra to plot the data 
plotRGB(kr_2024_rgb, r=1, g=2, b=3, stretch="lin")
plot(st_geometry(savanna_plots), add=TRUE, col="orange", pch=16)
plot(st_geometry(forest_plots), add=TRUE, col="darkgreen", pch=16)


```

### Transforming the data to map it on ggplot

```{r}
## Prep the data to plot it in ggplot
# define a single, shared CRS object
target_crs <- st_crs(32648) # UTM Zone 48

# make sure that the sf points are using the same crs
savanna_plots <- st_transform(savanna_plots, target_crs)
forest_plots <- st_transform(forest_plots, target_crs)

# Align the raster to the plots
all_plots <- rbind(savanna_plots, forest_plots) # Use original to avoid double-transforming
st_crs(all_plots) <- "EPSG:32648" 

# Force the raster to match the plot extent (Georeferencing)
plot_extent <- ext(all_plots) + 500
ext(kr_2024_rgb) <- plot_extent
crs(kr_2024_rgb) <- st_crs(all_plots)$wkt

# align the areas to the crop region
# Crop to the actual area of interest to keep the data frame small
kr_cropped <- crop(kr_2024_rgb, plot_extent)
kr_rescaled <- stretch(kr_cropped, minv = 0, maxv = 1)

# Convert to data frame for ggplot
kr_df <- as.data.frame(kr_rescaled, xy = TRUE, na.rm = TRUE)
colnames(kr_df)[3:5] <- c("R", "G", "B")

# ggplot
p1 <- ggplot() +
  geom_raster(data = kr_df, aes(x = x, y = y, fill = rgb(R, G, B))) +
  scale_fill_identity() +
  geom_sf(data = savanna_plots, fill = "orange", shape = 21, size = 3, color = "white") +
  geom_sf(data = forest_plots, fill = "darkgreen", shape = 21, size = 3, color = "white") +
  theme_minimal() +
  labs(title = "Field Sites: Cambodia (2024)",
       subtitle = "CRS: UTM Zone 48N")

p1
```

Enhance map

```{r}
library(ggspatial)

# Create the regional map
p2 <- ggplot() +
  # 1. Add a regional basemap (zooms out automatically)
  #annotation_map_tile(type = "osm", zoom = 10) + 
  
  # 2. Add your high-res raster (it will appear as a small patch)
  geom_raster(data = kr_df, aes(x = x, y = y, fill = rgb(R, G, B))) +
  scale_fill_identity() +
  
  # 3. Add your points
  geom_sf(data = savanna_plots, fill = "orange", shape = 21, size = 2) +
  geom_sf(data = forest_plots, fill = "darkgreen", shape = 21, size = 2) +
  # 4. Set the view to a larger area (e.g., Cambodia)
  # Adjust these coordinates to zoom out
  coord_sf(xlim = c(min(kr_df$x) - 100, max(kr_df$x) + 100),
                    ylim = c(min(kr_df$y) - 100, max(kr_df$y) + 100),
                    crs = st_crs(32648)) + 
  labs(title = "Field Sites: Cambodia (2024)",
       subtitle = "CRS: UTM Zone 48N")+
  theme_minimal()


p2


```

### Create a regional map

```{r}
# Trying a new method to zoom out of the site and look at the other provinces 
library(rnaturalearth)
library(ggspatial)

# Get the Cambodia border
cambodia <- ne_countries(scale = "medium", country = "cambodia", returnclass = "sf") %>%
  st_transform(32648)

# Plot with a Minimalist Type
p3 <- ggplot() +
  # Use a light, minimal basemap style
  annotation_map_tile(type = "cartolight", zoom = 7) + 
  
  # Draw the country outline
  geom_sf(data = cambodia, fill = NA, color = "black", size = 0.5) +
  
  # Add your sites (scaled up slightly so they are visible)
  geom_sf(data = savanna_plots, fill = "orange", shape = 21, size = 3, color = "white") +
  geom_sf(data = forest_plots, fill = "darkgreen", shape = 21, size = 3, color = "white") +
  
  # Minimal Compass and Scale
  annotation_north_arrow(location = "br", which_north = "true", 
                         style = north_arrow_minimal()) +
  annotation_scale(location = "bl", style = "ticks") +
  
  # Clean up the theme
  theme_void() +
  theme(panel.background = element_rect(fill = "white", color = NA)) +
  labs(title = "Project Sites: Cambodia")

p3

# using a different base layer to zoom out of the field site and look at the broader region

#provinces <- ne_states(country = "cambodia", returnclass = "sf") %>%
#st_transform(32648)
# isolate the specific province from the dataset so that it can be highlighted
#siem_reap <- provinces %>% filter(name == "Siemreab")

site_plot <- ggplot() +
  # Background Tiles (Minimal Carto Light)
  annotation_map_tile(type = "cartolight", 
                      zoom = 9) + 
   #targeted province
#  geom_sf(data = siem_reap, 
#          fill = NA, 
#          color = "black", 
#          size = 0.8) +

  # Field Sites with LEGEND mapping
  # Note: fill is moved inside aes() to create the legend
  geom_sf(data = savanna_plots, 
          aes(fill = "Savanna"), 
          shape = 21, 
          size = 2, 
          color = "white") +
  geom_sf(data = forest_plots, 
          aes(fill = "Forest"), 
          shape = 21, 
          size = 2, 
          color = "white") +
  # Define Legend Colors and Title
  scale_fill_manual(values = c("Savanna" = "orange", 
                               "Forest" = "darkgreen"), 
                    name = "Site Plot Type") +
coord_sf(xlim = c(103.25, 105.0), 
           ylim = c(12.91, 14.28),
           crs = st_crs(4326), # Using standard Lat/Lon for the zoom
           expand = FALSE)+
  theme_light(base_size = 14) + 
  theme(legend.position = "bottom", 
        plot.title = element_text(hjust = 0.25)) +
  labs(title = "Field Sites: Siem Reap Province, Cambodia")

site_plot

```

### Map for the poster

```{r}

## making a simple map with rnaturalearth
library(rnaturalearth)
library(rnaturalearthdata)

# add Cambodia's provincial boundaries
# returnclass = "sf" is critical for ggplot Compatibility
cambodia <- ne_states(country = "cambodia", returnclass = "sf")

# Download Lakes (Polygons)
lakes <- ne_download(scale = 50, 
                     type = "lakes", 
                     category = "physical", 
                     returnclass = "sf")

# Download Rivers (Lines)
rivers <- ne_download(scale = 50, 
                      type = "rivers_lake_centerlines", 
                      category = "physical", 
                      returnclass = "sf")
# add the capital city to the map
capital_data <- data.frame(
  city = "Phnom Penh",
  lat = 11.5564, 
  lon = 104.9282
)
# make an sf object 
capital_sf <- st_as_sf(capital_data, coords = c("lon", "lat"), crs = 4326)


# create a country map with field site and province
map1 <- ggplot(data = cambodia) +
  geom_sf(fill = "#a6dba0", 
          color = "grey25") +
  # Layer 2: Rivers (Blue lines)
  geom_sf(data = rivers, 
          color = "dodgerblue", 
          linewidth = 0.5) +
  # Layer 3: Lakes (Blue polygons)
  geom_sf(data = lakes, 
          fill = "dodgerblue", 
          color = "dodgerblue") +
  geom_sf(data = subset(cambodia, 
                        name_en == "Siem Reap"), 
          fill = "#7b3294", 
          alpha = 0.75, 
          color = "black") +
  geom_sf(data = savanna_plots, 
          pch = 4, 
          size = 2) +
  annotate("text", 
           x = 104.1, 
           y = 13.5, 
           label = "Siem Reap", 
           color = "black", 
           fontface = "bold", 
           size = 3) +
  geom_sf(data = capital_sf, color = "black", size = 3, shape = 18) +
  geom_sf_text(data = capital_sf, aes(label = city), nudge_y = 0.2, fontface = "bold", size = 2.5) +
  # CRITICAL: Zoom into Cambodia coordinates
  coord_sf(xlim = c(102, 108), 
           ylim = c(10, 15)) +
  theme_minimal() +
  theme(
    panel.grid = element_blank(),
    panel.border = element_rect(color = "black", fill = NA, size = 1),
    axis.text = element_text(size = 8)
  ) + labs(x = "Longitude", y = "Latitude")+
  labs(title = "Study site in Cambodia", subtitle = "Potential field site in Siem Reap Province")
map1

```

creating the Southeast Asia regional map

```{r}

sea_neighbors <- sea %>% filter(admin != "Cambodia")
cambodia_main <- ne_countries(country = "cambodia", returnclass = "sf", scale = "medium")

# calculate the center of the countries for placing 
# This creates X and Y columns for every country in your 'sea' object
neighbor_labels <- sea_neighbors %>%
  st_centroid() %>%
  st_coordinates() %>%
  as.data.frame() %>%
  mutate(name = sea_neighbors$name)

# 3. Create the Plot
inset_map2 <- ggplot() +
  # Draw neighbors in purple (no overlap now!)
  geom_sf(data = sea_neighbors, fill = "#c2a5cf", color = "#f7f7f7") +
  
  # Draw Cambodia in solid green (no alpha needed since there's no purple under it)
  geom_sf(data = cambodia_main, fill = "#a6dba0", color = "black") + 
  
  # AUTOMATED TEXT: Uses the coordinates we calculated
  geom_text(data = neighbor_labels, aes(x = X, y = Y, label = name), 
            size = 2.5, fontface = "bold", check_overlap = TRUE) +
  
  coord_sf(xlim = c(95, 115), ylim = c(5, 25)) + 
  theme_void() + 
  theme(panel.border = element_rect(color = "black", fill = NA)) +
  labs(title = "Southeast Asia", size = 2.75)

inset_map2


# draft final map with inset map
final_map2 <- map1 + 
  inset_element(
    inset_map, 
    left = 0.6,   # Horizontal start (0 to 1)
    bottom = 0.05, # Vertical start
    right = 0.99,  # Horizontal end
    top = 0.4,     # Vertical end
    align_to = 'panel',
    clip = FALSE   # Keeps the border from being cut off
  )

# View the result
final_map2

```
